---
title: "Clustering for Regime-Based Sector Rotation"
author: "Bao Nguyen"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-width: 12
    fig-height: 8
    smooth-scroll: true
execute:
  warning: false
  message: false
  echo: false
  cache: false
---

<!----------------------------------------------------------------------
  SETUP & DATA LOADING — all hidden computation
----------------------------------------------------------------------->

```{r setup}
#| output: false
library(tidyverse)
library(lubridate)
library(tidyquant)
library(tidymodels)
library(tidyclust)
library(timetk)
library(TTR)
library(PerformanceAnalytics)
library(xts)
library(zoo)
library(quantmod)
library(RTL)
library(lattice)
library(plotly)
library(foreach)
library(doParallel)
library(dlm)
library(cluster)
library(factoextra)
library(patchwork)
library(ggrepel)
library(gt)

# Helper: rolling function with partial window support
roll_partial <- function(x, width, FUN, min_obs = 5, ...) {
  rollapply(x, width = width, FUN = function(z) {
    z_clean <- z[!is.na(z)]
    if (length(z_clean) < min_obs) return(NA_real_)
    FUN(z_clean, ...)
  }, fill = NA, align = "right", partial = TRUE)
}

# Helper: Kalman filter for a single indicator series (local level model)
kf_level <- function(x, min_obs = 30) {
  out <- rep(NA_real_, length(x))
  if (sum(!is.na(x)) < min_obs) return(out)
  fit <- tryCatch(
    dlmMLE(x, parm = c(0, 0),
           build = function(p) dlmModPoly(1, dV = exp(p[1]), dW = exp(p[2]))),
    error = function(e) NULL
  )
  if (is.null(fit) || fit$convergence != 0) return(out)
  mod  <- dlmModPoly(1, dV = exp(fit$par[1]), dW = exp(fit$par[2]))
  filt <- dlmFilter(x, mod)
  drop(filt$m[-1])
}

# Helper: Kalman filter for prediction quality tracking (predict-correct)
kf_exposure <- function(obs, min_obs = 12) {
  n <- length(obs)
  out <- list(
    state      = rep(NA_real_, n),
    innovation = rep(NA_real_, n),
    V = NA_real_, W = NA_real_
  )
  valid <- !is.na(obs)
  if (sum(valid) < min_obs) return(out)

  fit <- tryCatch(
    dlmMLE(obs, parm = c(0, 0),
           build = function(p) dlmModPoly(1, dV = exp(p[1]), dW = exp(p[2]))),
    error = function(e) NULL
  )
  if (is.null(fit) || fit$convergence != 0) return(out)

  V <- exp(fit$par[1])
  W <- exp(fit$par[2])
  mod  <- dlmModPoly(1, dV = V, dW = W)
  filt <- dlmFilter(obs, mod)

  state <- drop(filt$m[-1])
  innov <- drop(obs - filt$f)

  out$state      <- state
  out$innovation <- innov
  out$V          <- V
  out$W          <- W
  out
}

# Helper: rolling beta (two-series covariance / variance)
roll_beta <- function(sector_ret, spy_ret, width = 252, min_obs = 20) {
  mat <- cbind(sector_ret, spy_ret)
  rollapply(mat, width = width, FUN = function(x) {
    cc <- complete.cases(x)
    if (sum(cc) < min_obs) return(NA_real_)
    cov(x[cc, 1], x[cc, 2]) / var(x[cc, 2])
  }, by.column = FALSE, fill = NA, align = "right", partial = TRUE)
}

# Helper: Omega ratio (L=0)
omega_fn <- function(rets, L = 0) {
  excess <- rets - L
  sum(pmax(excess, 0)) / sum(pmax(-excess, 0))
}

# Helper: portfolio weights from a position tibble
compute_weights <- function(pos_tbl) {
  pos_tbl %>%
    group_by(date) %>%
    mutate(
      n_long  = sum(signal == 1L,  na.rm = TRUE),
      n_short = sum(signal == -1L, na.rm = TRUE),
      weight  = case_when(
        signal ==  1L & n_long  > 0 ~  1 / n_long,
        signal == -1L & n_short > 0 ~ -1 / n_short,
        TRUE                        ~  0
      )
    ) %>%
    ungroup()
}
```

```{r params}
#| output: false
tickers <- c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
             "XLK", "XLP", "XLRE", "XLU", "XLV", "XLY")

etf_inception <- tribble(
  ~ticker, ~inception,
  "SPY",   "1993-01-29",
  "XLB",   "1998-12-22",
  "XLC",   "2018-06-18",
  "XLE",   "1998-12-22",
  "XLF",   "1998-12-22",
  "XLI",   "1998-12-22",
  "XLK",   "1998-12-22",
  "XLP",   "1998-12-22",
  "XLRE",  "2015-10-07",
  "XLU",   "1998-12-22",
  "XLV",   "1998-12-22",
  "XLY",   "1998-12-22"
) %>% mutate(inception = as.Date(inception))

TRAIN_START <- as.Date("2000-06-01")
TRAIN_END   <- as.Date("2024-12-31")
TEST_START  <- as.Date("2025-01-01")
TEST_END    <- as.Date("2026-03-31")
PULL_FROM   <- as.Date("1998-01-01")

sector_tickers <- setdiff(tickers, "SPY")
defensive_tickers <- c("XLU", "XLP", "XLV", "XLRE")
cyclical_tickers  <- c("XLK", "XLY", "XLF", "XLI", "XLB", "XLE", "XLC")
```

```{r load-data}
#| cache: true

load_etf <- function(ticker) {
  raw <- tryCatch(
    tidyquant::tq_get(ticker, from = format(PULL_FROM)),
    error = function(e) NULL
  )
  if (is.null(raw) || isFALSE(raw) || nrow(raw) == 0) {
    warning(paste("tq_get failed for:", ticker))
    return(NULL)
  }
  raw %>%
    dplyr::rename_all(tools::toTitleCase) %>%
    timetk::tk_xts(date_var = Date) %>%
    quantmod::adjustOHLC(use.Adjusted = TRUE) %>%
    timetk::tk_tbl(rename_index = "Date") %>%
    dplyr::select(-Adjusted) %>%
    dplyr::mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
    dplyr::rename(date = Date) %>%
    dplyr::mutate(ticker = ticker)
}

raw_data <- tickers %>%
  purrr::map(load_etf) %>%
  purrr::compact() %>%
  dplyr::bind_rows() %>%
  left_join(etf_inception, by = "ticker") %>%
  filter(date >= inception) %>%
  select(-inception) %>%
  arrange(ticker, date)

train_data <- raw_data %>% filter(date >= TRAIN_START, date <= TRAIN_END)
test_data  <- raw_data %>% filter(date >= TEST_START,  date <= TEST_END)
```

```{r load-rates}
#| cache: true

rates_raw <- bind_rows(
  tq_get("^TNX", from = format(PULL_FROM)) %>%
    transmute(date, series = "y10", value = close),
  tq_get("DGS2", get = "economic.data", from = format(PULL_FROM)) %>%
    transmute(date, series = "y2", value = price),
  tq_get("T10YIE", get = "economic.data", from = format(PULL_FROM)) %>%
    transmute(date, series = "breakeven", value = price)
) %>%
  pivot_wider(names_from = series, values_from = value)

rates <- rates_raw %>%
  mutate(yield_curve = y10 - y2) %>%
  filter(date >= TRAIN_START)

rf_daily <- tq_get("^IRX", from = format(TRAIN_START)) %>%
  transmute(date, rf_daily = close / 100 / 252)
```

```{r compute-features}
#| cache: true

vix_raw <- tidyquant::tq_get("^VIX", from = format(PULL_FROM)) %>%
  select(date, vix = close)

spy_returns <- raw_data %>%
  filter(ticker == "SPY", date >= TRAIN_START) %>%
  arrange(date) %>%
  mutate(spy_retClCl = Close / lag(Close) - 1,
         spy_log_ret = log(Close / lag(Close))) %>%
  select(date, spy_retClCl, spy_log_ret)

RVOL_WINDOW  <- 21
BETA_WINDOW  <- 252
ZSCORE_WINDOW <- 252

features_long <- raw_data %>%
  filter(date >= TRAIN_START) %>%
  group_by(ticker) %>%
  arrange(date) %>%
  mutate(
    retClCl     = Close / lag(Close) - 1,
    log_ret     = log(Close / lag(Close)),
    rvol_21     = roll_partial(retClCl, width = RVOL_WINDOW, FUN = sd,
                               min_obs = 5) * sqrt(252) * 100,
    log_ret_21  = roll_partial(log_ret, width = RVOL_WINDOW, FUN = sum,
                               min_obs = 1)
  ) %>%
  ungroup() %>%
  left_join(spy_returns, by = "date") %>%
  left_join(vix_raw, by = "date") %>%
  group_by(ticker) %>%
  arrange(date) %>%
  mutate(
    beta_252 = roll_beta(retClCl, spy_retClCl, width = BETA_WINDOW, min_obs = 20),
    spread   = rvol_21 - vix,
    spread_z = {
      sp_mean <- roll_partial(spread, width = ZSCORE_WINDOW, FUN = mean, min_obs = 30)
      sp_sd   <- roll_partial(spread, width = ZSCORE_WINDOW, FUN = sd,   min_obs = 30)
      (spread - sp_mean) / sp_sd
    }
  ) %>%
  ungroup() %>%
  select(date, ticker, retClCl, log_ret, rvol_21, log_ret_21,
         beta_252, vix, spread, spread_z)

spy_ref <- features_long %>%
  filter(ticker == "SPY") %>%
  select(date, spy_rvol_21 = rvol_21, spy_log_ret_21 = log_ret_21)

features_long <- features_long %>%
  left_join(spy_ref, by = "date") %>%
  mutate(
    rel_vol      = rvol_21 - spy_rvol_21,
    rel_momentum = log_ret_21 - spy_log_ret_21
  ) %>%
  select(-spy_rvol_21, -spy_log_ret_21)

features_wide <- features_long %>%
  select(date, ticker, rvol_21, spread, spread_z, log_ret_21, beta_252,
         rel_vol, rel_momentum) %>%
  pivot_wider(names_from = ticker,
              values_from = c(rvol_21, spread, spread_z, log_ret_21, beta_252,
                              rel_vol, rel_momentum),
              names_glue = "{ticker}_{.value}") %>%
  left_join(vix_raw, by = "date") %>%
  left_join(rates %>% select(date, yield_curve, y10, y2, breakeven), by = "date") %>%
  mutate(vrp = vix - SPY_rvol_21)
```

```{r kalman-filter}
kf_cols <- c("vix", "vrp", "yield_curve", "breakeven")

features_kf <- features_wide %>%
  arrange(date) %>%
  mutate(across(all_of(kf_cols), kf_level))
```

<!----------------------------------------------------------------------
  CLUSTERING COMPUTATION
----------------------------------------------------------------------->

```{r cluster-prep}
cluster_cols <- c("vix", "vrp", "yield_curve", "breakeven")

cluster_input <- features_kf %>%
  filter(date >= TRAIN_START, date <= TRAIN_END) %>%
  select(date, all_of(cluster_cols)) %>%
  drop_na()

cluster_recipe <- recipe(~ ., data = cluster_input %>% select(-date)) %>%
  step_normalize(all_predictors())

cluster_prep <- prep(cluster_recipe)
cluster_mat  <- bake(cluster_prep, new_data = NULL) %>% as.matrix()
```

```{r cluster-k-select}
set.seed(42)

sil_sample_idx <- sample(nrow(cluster_mat), min(500, nrow(cluster_mat)))
sil_sample_mat <- cluster_mat[sil_sample_idx, ]

k_metrics <- map_dfr(2:6, function(k) {
  km_full   <- kmeans(cluster_mat, centers = k, nstart = 25, iter.max = 200)
  km_sample <- kmeans(sil_sample_mat, centers = km_full$centers, nstart = 1, iter.max = 50)
  sil       <- cluster::silhouette(km_sample$cluster, dist(sil_sample_mat))
  tibble(k = k, wss = km_full$tot.withinss, silhouette = mean(sil[, 3]))
})

best_k_sil <- k_metrics$k[which.max(k_metrics$silhouette)]
```

```{r cluster-fit}
K_CLUSTERS <- 3
K_SECTORS  <- 2

set.seed(42)
km_fit_full <- kmeans(cluster_mat, centers = K_CLUSTERS, nstart = 25, iter.max = 200)

cluster_assignments <- cluster_input %>%
  mutate(km_cluster = km_fit_full$cluster)
```

```{r sector-k-prep}
# Sector daily returns + momentum + beta
sector_ret <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  select(date, ticker, retClCl, beta_252) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(momentum_21 = slider::slide_dbl(retClCl, ~prod(1 + .x, na.rm = TRUE) - 1,
                                         .before = 20, .complete = TRUE)) %>%
  ungroup()

# Month-end rebalancing dates
rebal_dates <- cluster_input %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym) %>%
  slice_max(date, n = 1) %>%
  ungroup() %>%
  arrange(date) %>%
  pull(date)
```

```{r sector-k-select}
sample_idx <- round(seq(max(1, length(rebal_dates) %/% 4),
                        length(rebal_dates),
                        length.out = min(20, length(rebal_dates))))
sample_rebal <- rebal_dates[sample_idx]

k2_metrics <- map_dfr(2:5, function(k) {
  per_date <- map_dfr(sample_rebal, function(t) {
    idx <- which(cluster_input$date == t)
    if (length(idx) == 0 || idx < 9) return(NULL)

    window_mat <- cluster_mat[1:idx, , drop = FALSE]
    set.seed(42 + which(rebal_dates == t))
    km_regime <- kmeans(window_mat, centers = 3, nstart = 25, iter.max = 200)
    current_regime <- km_regime$cluster[idx]
    regime_dates <- cluster_input$date[1:idx][km_regime$cluster == current_regime]
    regime_dates <- tail(regime_dates, 120)

    regime_stats <- sector_ret %>%
      filter(date %in% regime_dates, ticker %in% sector_tickers) %>%
      group_by(ticker) %>%
      summarise(regime_ret = mean(retClCl, na.rm = TRUE),
                regime_vol = sd(retClCl, na.rm = TRUE),
                regime_sharpe = if_else(regime_vol > 0, regime_ret / regime_vol, 0),
                .groups = "drop") %>%
      drop_na()

    beta_now <- sector_ret %>% filter(date == t) %>% select(ticker, beta_252)
    regime_stats <- regime_stats %>%
      left_join(beta_now, by = "ticker") %>%
      filter(!is.na(beta_252))

    if (nrow(regime_stats) < k + 1) return(NULL)

    feat <- regime_stats %>%
      select(regime_ret, regime_vol, regime_sharpe, beta_252) %>%
      scale()
    feat[is.nan(feat)] <- 0

    set.seed(100 + which(rebal_dates == t))
    km_s <- kmeans(feat, centers = k, nstart = 25, iter.max = 200)
    sil  <- cluster::silhouette(km_s$cluster, dist(feat))

    tibble(k = k, wss = km_s$tot.withinss, silhouette = mean(sil[, 3]))
  })

  per_date %>%
    summarise(k = first(k),
              wss = mean(wss, na.rm = TRUE),
              silhouette = mean(silhouette, na.rm = TRUE))
})

best_k2_sil <- k2_metrics$k[which.max(k2_metrics$silhouette)]
```

```{r sector-silhouette-data}
sil_sector_data <- map_dfr(sample_rebal, function(t) {
  idx <- which(cluster_input$date == t)
  if (length(idx) == 0 || idx < 9) return(NULL)

  window_mat <- cluster_mat[1:idx, , drop = FALSE]
  set.seed(42 + which(rebal_dates == t))
  km_regime <- kmeans(window_mat, centers = 3, nstart = 25, iter.max = 200)
  current_regime <- km_regime$cluster[idx]
  regime_dates <- cluster_input$date[1:idx][km_regime$cluster == current_regime]
  regime_dates <- tail(regime_dates, 120)

  regime_stats <- sector_ret %>%
    filter(date %in% regime_dates, ticker %in% sector_tickers) %>%
    group_by(ticker) %>%
    summarise(regime_ret = mean(retClCl, na.rm = TRUE),
              regime_vol = sd(retClCl, na.rm = TRUE),
              regime_sharpe = if_else(regime_vol > 0, regime_ret / regime_vol, 0),
              .groups = "drop") %>%
    drop_na()

  beta_now <- sector_ret %>% filter(date == t) %>% select(ticker, beta_252)
  regime_stats <- regime_stats %>%
    left_join(beta_now, by = "ticker") %>%
    filter(!is.na(beta_252))

  if (nrow(regime_stats) < 3) return(NULL)

  feat <- regime_stats %>%
    select(regime_ret, regime_vol, regime_sharpe, beta_252) %>%
    scale()
  feat[is.nan(feat)] <- 0

  set.seed(100 + which(rebal_dates == t))
  km_s <- kmeans(feat, centers = 2, nstart = 25, iter.max = 200)
  sil  <- cluster::silhouette(km_s$cluster, dist(feat))

  tibble(
    ticker    = regime_stats$ticker,
    cluster   = factor(km_s$cluster),
    sil_width = sil[, 3],
    snap_date = t
  )
})

sil_avg <- sil_sector_data %>%
  group_by(ticker) %>%
  summarise(
    sil_width = mean(sil_width, na.rm = TRUE),
    cluster   = names(sort(table(cluster), decreasing = TRUE))[1],
    .groups   = "drop"
  ) %>%
  mutate(cluster = factor(cluster)) %>%
  arrange(cluster, desc(sil_width)) %>%
  mutate(order = row_number())

avg_sil <- mean(sil_avg$sil_width)
```

```{r cluster-centroid-snapshots-data}
# Pick snapshot dates that are guaranteed to exist in rebal_dates
mid_date <- rebal_dates[length(rebal_dates) %/% 2]
end_date <- max(rebal_dates)
# First rebal_date in each regime cluster
regime_firsts <- cluster_assignments %>%
  filter(date %in% rebal_dates) %>%
  group_by(km_cluster) %>%
  slice_min(date, n = 1) %>%
  ungroup() %>%
  pull(date)
snapshot_dates <- sort(unique(c(regime_firsts, mid_date, end_date)))

snapshot_data <- map_dfr(snapshot_dates, function(t) {
  idx <- which(cluster_input$date == t)
  if (length(idx) == 0 || idx < K_CLUSTERS * 3) return(NULL)

  rebal_idx <- which(rebal_dates == t)
  if (length(rebal_idx) == 0) return(NULL)

  window_mat <- cluster_mat[1:idx, , drop = FALSE]
  set.seed(42L + rebal_idx[1])
  km_regime <- kmeans(window_mat, centers = K_CLUSTERS, nstart = 25, iter.max = 200)
  current_regime <- km_regime$cluster[idx]
  regime_dates <- cluster_input$date[1:idx][km_regime$cluster == current_regime]
  regime_dates <- tail(regime_dates, 120)

  regime_stats <- sector_ret %>%
    filter(date %in% regime_dates, ticker %in% sector_tickers) %>%
    group_by(ticker) %>%
    summarise(
      regime_ret    = mean(retClCl, na.rm = TRUE),
      regime_vol    = sd(retClCl, na.rm = TRUE),
      regime_sharpe = if_else(regime_vol > 0, regime_ret / regime_vol, 0),
      .groups = "drop"
    ) %>%
    drop_na()

  beta_now <- sector_ret %>% filter(date == t) %>% select(ticker, beta_252)
  regime_stats <- regime_stats %>%
    left_join(beta_now, by = "ticker") %>%
    filter(!is.na(beta_252))

  if (nrow(regime_stats) < K_SECTORS + 1) return(NULL)

  feat <- regime_stats %>%
    select(regime_ret, regime_vol, regime_sharpe, beta_252) %>%
    scale()
  feat[is.nan(feat)] <- 0

  set.seed(100L + rebal_idx[1])
  km_s <- kmeans(feat, centers = K_SECTORS, nstart = 25, iter.max = 200)

  centroids <- as_tibble(km_s$centers, .name_repair = "minimal")
  names(centroids) <- c("regime_ret", "regime_vol", "regime_sharpe", "beta_252")
  centroids$cluster <- factor(seq_len(K_SECTORS))

  regime_stats$sector_cluster <- km_s$cluster
  best_cl <- regime_stats %>%
    group_by(sector_cluster) %>%
    summarise(avg = mean(regime_ret), .groups = "drop") %>%
    slice_max(avg, n = 1, with_ties = FALSE) %>%
    pull(sector_cluster)

  centroids$is_best <- centroids$cluster == best_cl
  centroids$snapshot_date <- t
  centroids$regime <- current_regime
  centroids
})
```

<!----------------------------------------------------------------------
  SIGNAL GENERATION
----------------------------------------------------------------------->

```{r signal-params}
N_LONG          <- 3
N_SHORT         <- 0
VIX_THRESHOLD   <- 25
KF_HIGH_Q       <- 0.65
KF_LOW_Q        <- 0.20
MAX_REGIME_DAYS <- 120
```

```{r signal-generation}
two_layer_signal <- function(t, i, N_LONG, N_SHORT, VIX_THRESHOLD,
                              cluster_input_df, cluster_mat_full,
                              seed_offset = 0L, min_regime_obs = 20L) {

  flat <- tibble(rebal_date = t, regime = NA_integer_, sector_cluster = NA_integer_,
                 n_regime_dates = 0L, ticker = sector_tickers,
                 avg_ret = NA_real_, signal = 0L,
                 pred_spread = NA_real_, regime_dist = NA_real_)

  current_vix <- features_wide %>% filter(date == t) %>% pull(vix)
  if (length(current_vix) == 1 && !is.na(current_vix) && current_vix > VIX_THRESHOLD)
    return(flat)

  idx <- which(cluster_input_df$date == t)
  if (length(idx) == 0 || idx < K_CLUSTERS * 3) return(flat)

  window_mat <- cluster_mat_full[1:idx, , drop = FALSE]
  set.seed(42 + i + seed_offset)
  km_regime <- kmeans(window_mat, centers = K_CLUSTERS, nstart = 25, iter.max = 200)

  current_regime <- km_regime$cluster[idx]
  regime_dates   <- cluster_input_df$date[1:idx][km_regime$cluster == current_regime]
  regime_dist <- sqrt(sum((window_mat[idx, ] - km_regime$centers[current_regime, ])^2))

  if (length(regime_dates) < min_regime_obs) return(flat)

  regime_dates <- tail(regime_dates, MAX_REGIME_DAYS)

  regime_sector_stats <- sector_ret %>%
    filter(date %in% regime_dates, ticker %in% sector_tickers) %>%
    group_by(ticker) %>%
    summarise(
      regime_ret    = mean(retClCl, na.rm = TRUE),
      regime_vol    = sd(retClCl, na.rm = TRUE),
      regime_sharpe = if_else(regime_vol > 0, regime_ret / regime_vol, 0),
      .groups = "drop"
    ) %>%
    drop_na()

  beta_now <- sector_ret %>%
    filter(date == t) %>%
    select(ticker, beta_252)

  regime_sector_stats <- regime_sector_stats %>%
    left_join(beta_now, by = "ticker") %>%
    filter(!is.na(beta_252))

  if (nrow(regime_sector_stats) < K_SECTORS + 1) return(flat)

  feat_sector <- regime_sector_stats %>%
    select(regime_ret, regime_vol, regime_sharpe, beta_252) %>%
    scale()

  feat_sector[is.nan(feat_sector)] <- 0
  if (any(!is.finite(feat_sector))) return(flat)

  set.seed(100 + i + seed_offset)
  km_sector <- kmeans(feat_sector, centers = K_SECTORS, nstart = 25, iter.max = 200)

  regime_sector_stats$sector_cluster <- km_sector$cluster

  best_cl <- regime_sector_stats %>%
    group_by(sector_cluster) %>%
    summarise(avg = mean(regime_ret, na.rm = TRUE), .groups = "drop") %>%
    slice_max(avg, n = 1, with_ties = FALSE) %>%
    pull(sector_cluster)

  mom_now <- sector_ret %>%
    filter(date == t) %>%
    select(ticker, momentum_21)

  best_sectors <- regime_sector_stats %>%
    filter(sector_cluster == best_cl) %>%
    left_join(mom_now, by = "ticker") %>%
    arrange(desc(momentum_21)) %>%
    mutate(rank = row_number(),
           signal = if_else(rank <= N_LONG, 1L, 0L))

  all_sectors <- regime_sector_stats %>%
    select(ticker, sector_cluster, avg_ret = regime_ret) %>%
    left_join(best_sectors %>% select(ticker, signal), by = "ticker") %>%
    mutate(signal = replace_na(signal, 0L)) %>%
    {bind_rows(., tibble(
        ticker = setdiff(sector_tickers, .$ticker),
        sector_cluster = NA_integer_, avg_ret = NA_real_, signal = 0L))} %>%
    mutate(rebal_date = t, regime = current_regime,
           n_regime_dates = length(regime_dates),
           regime_dist = regime_dist)

  long_avg    <- mean(all_sectors$avg_ret[all_sectors$signal == 1L], na.rm = TRUE)
  nonlong_avg <- mean(all_sectors$avg_ret[all_sectors$signal == 0L], na.rm = TRUE)
  all_sectors$pred_spread <- if_else(
    is.finite(long_avg) & is.finite(nonlong_avg),
    long_avg - nonlong_avg, NA_real_
  )

  all_sectors
}

signals_raw <- map_dfr(seq_along(rebal_dates), function(i) {
  two_layer_signal(rebal_dates[i], i, N_LONG, N_SHORT, VIX_THRESHOLD,
                    cluster_input, cluster_mat)
})
```

<!----------------------------------------------------------------------
  TRAINING PERIOD P&L
----------------------------------------------------------------------->

```{r training-pnl}
spy_dates <- raw_data %>%
  filter(ticker == "SPY", date >= TRAIN_START, date <= TRAIN_END) %>%
  arrange(date) %>%
  pull(date)

exec_map <- tibble(rebal_date = rebal_dates) %>%
  rowwise() %>%
  mutate(exec_date = {
    candidates <- spy_dates[spy_dates > rebal_date]
    if (length(candidates) == 0) NA_Date_ else candidates[1]
  }) %>%
  ungroup() %>%
  filter(!is.na(exec_date))

daily_signals <- signals_raw %>%
  select(rebal_date, ticker, signal) %>%
  inner_join(exec_map, by = "rebal_date") %>%
  select(date = exec_date, ticker, signal)

all_combos <- expand_grid(date = spy_dates, ticker = sector_tickers)

daily_pos <- all_combos %>%
  left_join(daily_signals, by = c("date", "ticker")) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  fill(signal, .direction = "down") %>%
  mutate(signal = replace_na(signal, 0L)) %>%
  ungroup()

exec_dates_set <- unique(exec_map$exec_date)

daily_rets <- raw_data %>%
  filter(ticker %in% sector_tickers, date %in% spy_dates) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(
    ret_cc  = Close / dplyr::lag(Close) - 1,
    ret_oc  = Close / Open - 1,
    is_exec = date %in% exec_dates_set,
    ret_cc  = if_else(is_exec, ret_oc, ret_cc)
  ) %>%
  ungroup() %>%
  select(date, ticker, ret_cc)

TC_BPS <- 0.0005

rebal_costs <- compute_weights(daily_pos) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(prev_weight = dplyr::lag(weight, default = 0)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(
    tc = sum(abs(weight - prev_weight), na.rm = TRUE) * TC_BPS,
    .groups = "drop"
  )

daily_pnl <- compute_weights(daily_pos) %>%
  left_join(daily_rets, by = c("date", "ticker")) %>%
  group_by(date) %>%
  summarise(
    portfolio_ret_gross = sum(weight * replace_na(ret_cc, 0), na.rm = TRUE),
    n_long              = first(n_long),
    .groups             = "drop"
  ) %>%
  left_join(rebal_costs, by = "date") %>%
  mutate(
    tc            = replace_na(tc, 0),
    portfolio_ret = portfolio_ret_gross - tc,
    cumeq         = cumprod(1 + portfolio_ret),
    cumeq_gross   = cumprod(1 + portfolio_ret_gross)
  )

# KF predict-correct scaling
monthly_spread <- daily_pos %>%
  left_join(daily_rets, by = c("date", "ticker")) %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym, signal) %>%
  summarise(avg_ret = mean(ret_cc, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = signal, values_from = avg_ret,
              names_prefix = "sig_") %>%
  arrange(ym) %>%
  mutate(realized_spread = replace_na(sig_1, 0) - replace_na(sig_0, 0)) %>%
  pull(realized_spread)

kf_train <- kf_exposure(monthly_spread, min_obs = 12)

kf_state_valid <- kf_train$state[!is.na(kf_train$state)]
KF_HIGH_ABS <- quantile(kf_state_valid, KF_HIGH_Q, na.rm = TRUE)
KF_LOW_ABS  <- quantile(kf_state_valid, KF_LOW_Q,  na.rm = TRUE)

month_ends <- daily_pnl %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym) %>%
  slice_max(date, n = 1) %>%
  ungroup() %>%
  arrange(date) %>%
  pull(ym)

kf_month_scale <- tibble(
  ym         = month_ends,
  kf_state   = kf_train$state[seq_along(month_ends)],
  kf_innov   = kf_train$innovation[seq_along(month_ends)]
) %>%
  mutate(
    kf_scale = case_when(
      is.na(kf_state)          ~ 1.0,
      kf_state >= KF_HIGH_ABS  ~ 1.0,
      kf_state >= KF_LOW_ABS   ~ 0.5,
      TRUE                     ~ 0.0
    ),
    kf_scale = case_when(
      !is.na(kf_innov) & kf_innov > abs(KF_HIGH_ABS) ~ pmin(kf_scale + 0.5, 1.0),
      !is.na(kf_innov) & kf_innov < -abs(KF_HIGH_ABS) ~ pmax(kf_scale - 0.5, 0.0),
      TRUE ~ kf_scale
    ),
    next_ym_scale = dplyr::lead(kf_scale, default = 1.0)
  ) %>%
  mutate(ym = floor_date(ym %m+% months(1), "month")) %>%
  select(ym, next_ym_scale)

daily_pos_kf <- daily_pos %>%
  mutate(ym = floor_date(date, "month")) %>%
  left_join(kf_month_scale, by = "ym") %>%
  mutate(next_ym_scale = replace_na(next_ym_scale, 1.0)) %>%
  select(-ym)

rebal_costs_kf <- compute_weights(daily_pos_kf) %>%
  mutate(weight_s = weight * next_ym_scale) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(prev_ws = dplyr::lag(weight_s, default = 0)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(tc = sum(abs(weight_s - prev_ws), na.rm = TRUE) * TC_BPS, .groups = "drop")

daily_pnl_dd <- compute_weights(daily_pos_kf) %>%
  mutate(weight_s = weight * next_ym_scale) %>%
  left_join(daily_rets, by = c("date", "ticker")) %>%
  group_by(date) %>%
  summarise(
    ret_sectors = sum(weight_s * replace_na(ret_cc, 0), na.rm = TRUE),
    kf_scale    = first(next_ym_scale),
    n_long      = first(n_long),
    .groups     = "drop"
  ) %>%
  left_join(rebal_costs_kf, by = "date") %>%
  left_join(rf_daily, by = "date") %>%
  arrange(date) %>%
  mutate(
    tc       = replace_na(tc, 0),
    rf_daily = replace_na(rf_daily, 0),
    portfolio_ret_gross = ret_sectors + (1 - kf_scale) * rf_daily,
    portfolio_ret = portfolio_ret_gross - tc,
    cumeq_kf  = cumprod(1 + portfolio_ret)
  ) %>%
  select(date, cumeq_kf, portfolio_ret_kf = portfolio_ret, kf_scale)

# SPY buy-and-hold
spy_bh <- raw_data %>%
  filter(ticker == "SPY", date %in% spy_dates) %>%
  arrange(date) %>%
  mutate(
    ret_cc    = Close / dplyr::lag(Close) - 1,
    cumeq_spy = cumprod(1 + replace_na(ret_cc, 0))
  ) %>%
  select(date, cumeq_spy, ret_cc_spy = ret_cc)

# Benchmark strategies
ew_benchmark <- function(tickers, dates, raw) {
  raw %>%
    filter(ticker %in% tickers, date %in% dates) %>%
    arrange(ticker, date) %>%
    group_by(ticker) %>%
    mutate(ret = Close / dplyr::lag(Close) - 1) %>%
    ungroup() %>%
    group_by(date) %>%
    summarise(ret = mean(ret, na.rm = TRUE), .groups = "drop") %>%
    arrange(date) %>%
    mutate(cumeq = cumprod(1 + replace_na(ret, 0)))
}

spy_top3_bench <- function(dates, rebal_dates, raw, sector_ret_df, sector_ticks) {
  top3_signals <- map_dfr(rebal_dates, function(t) {
    mom <- sector_ret_df %>%
      filter(date == t, ticker %in% sector_ticks) %>%
      arrange(desc(momentum_21)) %>%
      slice_head(n = 3)
    if (nrow(mom) == 0) return(NULL)
    tibble(rebal_date = t, ticker = mom$ticker)
  })

  exec_map_bench <- tibble(rebal_date = rebal_dates) %>%
    rowwise() %>%
    mutate(exec_date = {
      cands <- dates[dates > rebal_date]
      if (length(cands) == 0) NA_Date_ else cands[1]
    }) %>%
    ungroup() %>% filter(!is.na(exec_date))

  daily_sig <- top3_signals %>%
    inner_join(exec_map_bench, by = "rebal_date") %>%
    select(date = exec_date, ticker)

  spy_ret <- raw %>%
    filter(ticker == "SPY", date %in% dates) %>%
    arrange(date) %>%
    mutate(ret_spy = Close / dplyr::lag(Close) - 1) %>%
    select(date, ret_spy)

  sector_daily <- raw %>%
    filter(ticker %in% sector_ticks, date %in% dates) %>%
    arrange(ticker, date) %>%
    group_by(ticker) %>%
    mutate(ret = Close / dplyr::lag(Close) - 1) %>%
    ungroup() %>%
    select(date, ticker, ret)

  all_dates_ticks <- expand_grid(date = dates, ticker = sector_ticks) %>%
    left_join(daily_sig %>% mutate(active = TRUE), by = c("date", "ticker")) %>%
    arrange(ticker, date) %>%
    group_by(ticker) %>%
    fill(active, .direction = "down") %>%
    mutate(active = replace_na(active, FALSE)) %>%
    ungroup()

  top3_ret <- all_dates_ticks %>%
    filter(active) %>%
    left_join(sector_daily, by = c("date", "ticker")) %>%
    group_by(date) %>%
    summarise(ret_top3 = mean(ret, na.rm = TRUE), .groups = "drop")

  spy_ret %>%
    left_join(top3_ret, by = "date") %>%
    arrange(date) %>%
    mutate(
      ret_top3 = replace_na(ret_top3, 0),
      ret_spy  = replace_na(ret_spy, 0),
      ret      = 0.5 * ret_spy + 0.5 * ret_top3,
      cumeq    = cumprod(1 + ret)
    )
}

bench_def  <- ew_benchmark(defensive_tickers, spy_dates, raw_data)
bench_cyc  <- ew_benchmark(cyclical_tickers,  spy_dates, raw_data)
bench_5050 <- spy_top3_bench(spy_dates, rebal_dates, raw_data, sector_ret, sector_tickers)

training_perf <- daily_pnl %>%
  left_join(spy_bh,      by = "date") %>%
  left_join(daily_pnl_dd, by = "date") %>%
  left_join(bench_def  %>% select(date, cumeq_def = cumeq), by = "date") %>%
  left_join(bench_cyc  %>% select(date, cumeq_cyc = cumeq), by = "date") %>%
  left_join(bench_5050 %>% select(date, cumeq_5050 = cumeq), by = "date")
```

```{r training-stats-compute}
n_days   <- nrow(training_perf)
n_years  <- n_days / 252

strat_rets       <- training_perf$portfolio_ret
strat_rets_gross <- training_perf$portfolio_ret_gross
spy_rets   <- replace_na(training_perf$ret_cc_spy, 0)

rf_joined <- training_perf %>%
  left_join(rf_daily, by = "date") %>%
  mutate(rf_daily = replace_na(rf_daily, 0))

excess_strat <- strat_rets    - rf_joined$rf_daily
excess_spy   <- spy_rets      - rf_joined$rf_daily

sharpe_strat <- (mean(excess_strat, na.rm = TRUE) /
                 sd(excess_strat,   na.rm = TRUE)) * sqrt(252)
sharpe_spy   <- (mean(excess_spy,   na.rm = TRUE) /
                 sd(excess_spy,     na.rm = TRUE)) * sqrt(252)

total_strat        <- last(training_perf$cumeq)       - 1
total_strat_gross  <- last(training_perf$cumeq_gross) - 1
total_spy          <- last(training_perf$cumeq_spy)   - 1
cagr_strat         <- (1 + total_strat)^(1/n_years)       - 1
cagr_strat_gross   <- (1 + total_strat_gross)^(1/n_years) - 1
cagr_spy           <- (1 + total_spy)^(1/n_years)         - 1
ann_vol_strat      <- sd(strat_rets,       na.rm = TRUE) * sqrt(252)
ann_vol_strat_gross<- sd(strat_rets_gross, na.rm = TRUE) * sqrt(252)
ann_vol_spy        <- sd(spy_rets,         na.rm = TRUE) * sqrt(252)

peak_strat        <- cummax(training_perf$cumeq)
peak_strat_gross  <- cummax(training_perf$cumeq_gross)
peak_spy          <- cummax(training_perf$cumeq_spy)
peak_kf_stop      <- cummax(replace_na(training_perf$cumeq_kf, 1))
max_dd_strat      <- min((training_perf$cumeq       - peak_strat)       / peak_strat,       na.rm = TRUE)
max_dd_strat_gross<- min((training_perf$cumeq_gross - peak_strat_gross) / peak_strat_gross, na.rm = TRUE)
max_dd_spy        <- min((training_perf$cumeq_spy   - peak_spy)         / peak_spy,         na.rm = TRUE)
max_dd_kf         <- min((replace_na(training_perf$cumeq_kf, 1) - peak_kf_stop) / peak_kf_stop, na.rm = TRUE)

mdd_date_strat   <- training_perf$date[which.min((training_perf$cumeq - peak_strat) / peak_strat)]
mdd_date_spy     <- training_perf$date[which.min((training_perf$cumeq_spy - peak_spy) / peak_spy)]
mdd_date_kf      <- training_perf$date[which.min((replace_na(training_perf$cumeq_kf, 1) - peak_kf_stop) / peak_kf_stop)]

excess_strat_gross <- strat_rets_gross - rf_joined$rf_daily
sharpe_strat_gross <- (mean(excess_strat_gross, na.rm = TRUE) /
                       sd(excess_strat_gross,   na.rm = TRUE)) * sqrt(252)

kf_rets   <- replace_na(training_perf$portfolio_ret_kf, 0)
excess_kf <- kf_rets - rf_joined$rf_daily
sharpe_kf <- (mean(excess_kf, na.rm = TRUE) / sd(excess_kf, na.rm = TRUE)) * sqrt(252)

omega_strat       <- omega_fn(strat_rets)
omega_strat_gross <- omega_fn(strat_rets_gross)
omega_spy         <- omega_fn(spy_rets)
omega_kf          <- omega_fn(kf_rets)
```

<!----------------------------------------------------------------------
  OPTIMIZATION
----------------------------------------------------------------------->

```{r opt-precompute}
#| cache: true

km_precomputed <- lapply(seq_along(rebal_dates), function(i) {
  t   <- rebal_dates[i]
  idx <- which(cluster_input$date == t)
  if (length(idx) == 0 || idx < K_CLUSTERS * 3) return(NULL)

  window_mat <- cluster_mat[1:idx, , drop = FALSE]
  set.seed(42 + i)
  km <- kmeans(window_mat, centers = K_CLUSTERS, nstart = 25, iter.max = 200)

  current_regime <- km$cluster[idx]
  vix_val <- features_wide %>% filter(date == t) %>% pull(vix)

  list(
    t             = t,
    regime        = current_regime,
    regime_dates  = cluster_input$date[1:idx][km$cluster == current_regime],
    current_vix   = if (length(vix_val) == 1) vix_val else NA_real_,
    regime_dist   = sqrt(sum((window_mat[idx, ] - km$centers[current_regime, ])^2))
  )
})
```

```{r opt-functions}
opt_weights <- function(pos_tbl) {
  pos_tbl %>%
    group_by(date) %>%
    mutate(
      n_long  = sum(signal == 1L,  na.rm = TRUE),
      n_short = sum(signal == -1L, na.rm = TRUE),
      weight  = case_when(
        signal ==  1L & n_long  > 0 ~  1 / n_long,
        signal == -1L & n_short > 0 ~ -1 / n_short,
        TRUE                        ~  0
      )
    ) %>%
    ungroup()
}

opt_signals <- function(km_results, N_LONG, N_SHORT = 0L,
                         VIX_THRESHOLD, sector_ret, sector_tickers,
                         K_SECTORS, min_regime_obs = 20L) {
  map_dfr(seq_along(km_results), function(i) {
    res <- km_results[[i]]
    if (is.null(res)) return(NULL)
    t <- res$t

    flat <- tibble(rebal_date = t, regime = res$regime, sector_cluster = NA_integer_,
                   n_regime_dates = length(res$regime_dates),
                   ticker = sector_tickers, avg_ret = NA_real_, signal = 0L,
                   pred_spread = NA_real_, regime_dist = NA_real_)

    if (!is.na(res$current_vix) && res$current_vix > VIX_THRESHOLD) return(flat)
    if (length(res$regime_dates) < min_regime_obs)                   return(flat)

    regime_dates_recent <- tail(res$regime_dates, MAX_REGIME_DAYS)

    regime_stats <- sector_ret %>%
      filter(date %in% regime_dates_recent, ticker %in% sector_tickers) %>%
      group_by(ticker) %>%
      summarise(regime_ret = mean(retClCl, na.rm = TRUE),
                regime_vol = sd(retClCl, na.rm = TRUE),
                regime_sharpe = if_else(regime_vol > 0, regime_ret / regime_vol, 0),
                .groups = "drop") %>%
      drop_na()

    beta_now <- sector_ret %>%
      filter(date == t) %>%
      select(ticker, beta_252)

    regime_stats <- regime_stats %>%
      left_join(beta_now, by = "ticker") %>%
      filter(!is.na(beta_252))

    if (nrow(regime_stats) < K_SECTORS + 1) return(flat)

    feat <- regime_stats %>% select(regime_ret, regime_vol, regime_sharpe, beta_252) %>% scale()
    feat[is.nan(feat)] <- 0
    if (any(!is.finite(feat))) return(flat)
    set.seed(100 + i)
    km_s <- kmeans(feat, centers = K_SECTORS, nstart = 25, iter.max = 200)
    regime_stats$sector_cluster <- km_s$cluster

    best_cl <- regime_stats %>%
      group_by(sector_cluster) %>%
      summarise(avg = mean(regime_ret), .groups = "drop") %>%
      slice_max(avg, n = 1, with_ties = FALSE) %>% pull(sector_cluster)

    mom_now <- sector_ret %>% filter(date == t) %>% select(ticker, momentum_21)

    best <- regime_stats %>%
      filter(sector_cluster == best_cl) %>%
      left_join(mom_now, by = "ticker") %>%
      arrange(desc(momentum_21)) %>%
      mutate(rank = row_number(), signal = if_else(rank <= N_LONG, 1L, 0L))

    all_sec <- regime_stats %>%
      select(ticker, sector_cluster, avg_ret = regime_ret) %>%
      left_join(best %>% select(ticker, signal), by = "ticker") %>%
      mutate(signal = replace_na(signal, 0L)) %>%
      {bind_rows(., tibble(ticker = setdiff(sector_tickers, .$ticker),
                            sector_cluster = NA_integer_, avg_ret = NA_real_, signal = 0L))} %>%
      mutate(rebal_date = t, regime = res$regime,
             n_regime_dates = length(res$regime_dates),
             regime_dist = res$regime_dist)

    long_avg    <- mean(all_sec$avg_ret[all_sec$signal == 1L], na.rm = TRUE)
    nonlong_avg <- mean(all_sec$avg_ret[all_sec$signal == 0L], na.rm = TRUE)
    all_sec$pred_spread <- if_else(
      is.finite(long_avg) & is.finite(nonlong_avg),
      long_avg - nonlong_avg, NA_real_
    )
    all_sec
  })
}

opt_metrics <- function(signals_raw, KF_HIGH_Q, KF_LOW_Q,
                         spy_dates, exec_map, sector_tickers,
                         daily_rets, rf_daily, TC_BPS = 0.0005) {

  if (is.null(signals_raw) || nrow(signals_raw) == 0)
    return(tibble(cagr=NA, sharpe=NA, omega=NA, max_dd=NA, total_ret=NA))

  daily_signals <- signals_raw %>%
    select(rebal_date, ticker, signal) %>%
    inner_join(exec_map, by = "rebal_date") %>%
    select(date = exec_date, ticker, signal)

  daily_pos <- expand_grid(date = spy_dates, ticker = sector_tickers) %>%
    left_join(daily_signals, by = c("date", "ticker")) %>%
    arrange(ticker, date) %>%
    group_by(ticker) %>%
    fill(signal, .direction = "down") %>%
    mutate(signal = replace_na(signal, 0L)) %>%
    ungroup()

  w1 <- opt_weights(daily_pos)

  tc1 <- w1 %>%
    arrange(ticker, date) %>%
    group_by(ticker) %>%
    mutate(prev_w = dplyr::lag(weight, default = 0)) %>%
    ungroup() %>%
    group_by(date) %>%
    summarise(tc = sum(abs(weight - prev_w), na.rm = TRUE) * TC_BPS, .groups = "drop")

  pnl1 <- w1 %>%
    left_join(daily_rets, by = c("date", "ticker")) %>%
    group_by(date) %>%
    summarise(ret_gross = sum(weight * replace_na(ret_cc, 0), na.rm = TRUE),
              n_long = first(n_long), .groups = "drop") %>%
    left_join(tc1, by = "date") %>%
    mutate(tc = replace_na(tc, 0),
           ret  = ret_gross - tc,
           cumeq = cumprod(1 + ret))

  m_spread <- daily_pos %>%
    left_join(daily_rets, by = c("date", "ticker")) %>%
    mutate(ym = floor_date(date, "month")) %>%
    group_by(ym, signal) %>%
    summarise(avg_ret = mean(ret_cc, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = signal, values_from = avg_ret,
                names_prefix = "sig_") %>%
    arrange(ym) %>%
    mutate(realized_spread = replace_na(sig_1, 0) - replace_na(sig_0, 0)) %>%
    pull(realized_spread)

  kf_res <- kf_exposure(m_spread, min_obs = 12)

  kf_valid <- kf_res$state[!is.na(kf_res$state)]
  if (length(kf_valid) < 2) {
    kf_hi <- 0; kf_lo <- 0
  } else {
    kf_hi <- quantile(kf_valid, KF_HIGH_Q, na.rm = TRUE)
    kf_lo <- quantile(kf_valid, KF_LOW_Q,  na.rm = TRUE)
  }

  m_ends <- pnl1 %>%
    mutate(ym = floor_date(date, "month")) %>%
    group_by(ym) %>% slice_max(date, n = 1) %>% ungroup() %>%
    arrange(date) %>% pull(ym)

  kf_sc <- tibble(
    ym = m_ends,
    kf_state = kf_res$state[seq_along(m_ends)],
    kf_innov = kf_res$innovation[seq_along(m_ends)]
  ) %>%
    mutate(
      kf_scale = case_when(
        is.na(kf_state)        ~ 1.0,
        kf_state >= kf_hi      ~ 1.0,
        kf_state >= kf_lo      ~ 0.5,
        TRUE                   ~ 0.0
      ),
      kf_scale = case_when(
        !is.na(kf_innov) & kf_innov >  abs(kf_hi) ~ pmin(kf_scale + 0.5, 1.0),
        !is.na(kf_innov) & kf_innov < -abs(kf_hi) ~ pmax(kf_scale - 0.5, 0.0),
        TRUE ~ kf_scale
      ),
      next_ym_scale = dplyr::lead(kf_scale, default = 1.0)
    ) %>%
    mutate(ym = floor_date(ym %m+% months(1), "month")) %>%
    select(ym, next_ym_scale)

  daily_pos_kf <- daily_pos %>%
    mutate(ym = floor_date(date, "month")) %>%
    left_join(kf_sc, by = "ym") %>%
    mutate(next_ym_scale = replace_na(next_ym_scale, 1.0)) %>%
    select(-ym)

  w2 <- opt_weights(daily_pos_kf) %>%
    mutate(weight_s = weight * next_ym_scale)

  tc2 <- w2 %>%
    arrange(ticker, date) %>%
    group_by(ticker) %>%
    mutate(prev_ws = dplyr::lag(weight_s, default = 0)) %>%
    ungroup() %>%
    group_by(date) %>%
    summarise(tc = sum(abs(weight_s - prev_ws), na.rm = TRUE) * TC_BPS, .groups = "drop")

  pnl2 <- w2 %>%
    left_join(daily_rets, by = c("date", "ticker")) %>%
    group_by(date) %>%
    summarise(ret_sectors = sum(weight_s * replace_na(ret_cc, 0), na.rm = TRUE),
              kf_scale = first(next_ym_scale),
              .groups = "drop") %>%
    left_join(tc2,     by = "date") %>%
    left_join(rf_daily, by = "date") %>%
    arrange(date) %>%
    mutate(
      tc       = replace_na(tc, 0),
      rf_daily = replace_na(rf_daily, 0),
      ret_gross = ret_sectors + (1 - kf_scale) * rf_daily,
      ret       = ret_gross - tc,
      cumeq     = cumprod(1 + ret)
    )

  n_years  <- nrow(pnl2) / 252
  rets     <- pnl2$ret
  rf       <- pnl2$rf_daily
  excess   <- rets - rf
  peak     <- cummax(pnl2$cumeq)
  max_dd   <- min((pnl2$cumeq - peak) / peak, na.rm = TRUE)
  total    <- last(pnl2$cumeq) - 1

  tibble(
    total_ret = total,
    cagr      = (1 + total)^(1 / n_years) - 1,
    sharpe    = (mean(excess, na.rm = TRUE) / sd(excess, na.rm = TRUE)) * sqrt(252),
    omega     = sum(pmax(rets, 0)) / sum(pmax(-rets, 0)),
    max_dd    = max_dd
  )
}
```

```{r opt-run}
param_grid <- expand_grid(
  KF_HIGH_Q     = seq(0.40, 0.85, by = 0.05),
  KF_LOW_Q      = seq(0.05, 0.40, by = 0.05)
) %>%
  filter(KF_HIGH_Q > KF_LOW_Q)

n_cores <- max(1, detectCores() - 1)
cl      <- makeCluster(n_cores)
registerDoParallel(cl)

clusterExport(cl, varlist = c(
  "km_precomputed", "K_CLUSTERS", "K_SECTORS", "MAX_REGIME_DAYS",
  "sector_ret", "sector_tickers",
  "daily_rets", "spy_dates", "exec_map", "rf_daily",
  "opt_weights", "opt_signals", "opt_metrics", "kf_exposure"
))
opt_sigs_fixed <- opt_signals(
  km_results     = km_precomputed,
  N_LONG         = N_LONG,
  VIX_THRESHOLD  = VIX_THRESHOLD,
  sector_ret     = sector_ret,
  sector_tickers = sector_tickers,
  K_SECTORS      = K_SECTORS
)

clusterExport(cl, "opt_sigs_fixed")

opt_results <- foreach(
  i         = seq_len(nrow(param_grid)),
  .combine  = bind_rows,
  .packages = c("tidyverse", "lubridate", "dlm")
) %dopar% {
  p <- param_grid[i, ]

  metrics <- opt_metrics(
    signals_raw    = opt_sigs_fixed,
    KF_HIGH_Q      = p$KF_HIGH_Q,
    KF_LOW_Q       = p$KF_LOW_Q,
    spy_dates      = spy_dates,
    exec_map       = exec_map,
    sector_tickers = sector_tickers,
    daily_rets     = daily_rets,
    rf_daily       = rf_daily
  )

  bind_cols(p, metrics)
}

stopCluster(cl)
registerDoSEQ()
```

```{r opt-best}
best <- opt_results %>% slice_max(omega, n = 1, with_ties = FALSE)

KF_HIGH_Q     <- best$KF_HIGH_Q
KF_LOW_Q      <- best$KF_LOW_Q
```

<!----------------------------------------------------------------------
  TEST PERIOD
----------------------------------------------------------------------->

```{r test-setup}
test_cluster_input <- features_kf %>%
  filter(date >= TEST_START, date <= TEST_END) %>%
  select(date, all_of(cluster_cols)) %>%
  drop_na()

test_cluster_mat <- bake(cluster_prep,
                         new_data = test_cluster_input %>% select(-date)) %>%
  as.matrix()

full_cluster_input <- bind_rows(cluster_input, test_cluster_input)
full_cluster_mat   <- rbind(cluster_mat, test_cluster_mat)

test_rebal_dates <- test_cluster_input %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym) %>%
  slice_max(date, n = 1) %>%
  ungroup() %>%
  arrange(date) %>%
  pull(date)

test_dates <- raw_data %>%
  filter(ticker == "SPY", date >= TEST_START, date <= TEST_END) %>%
  arrange(date) %>%
  pull(date)
```

```{r test-signals}
test_signals <- map_dfr(seq_along(test_rebal_dates), function(i) {
  two_layer_signal(test_rebal_dates[i], i, N_LONG, N_SHORT, VIX_THRESHOLD,
                    full_cluster_input, full_cluster_mat,
                    seed_offset = 1000L)
})
```

```{r test-pnl}
test_exec_map <- tibble(rebal_date = test_rebal_dates) %>%
  rowwise() %>%
  mutate(exec_date = {
    cands <- test_dates[test_dates > rebal_date]
    if (length(cands) == 0) NA_Date_ else cands[1]
  }) %>%
  ungroup() %>%
  filter(!is.na(exec_date))

test_daily_signals <- test_signals %>%
  select(rebal_date, ticker, signal) %>%
  inner_join(test_exec_map, by = "rebal_date") %>%
  select(date = exec_date, ticker, signal)

test_daily_pos <- expand_grid(date = test_dates, ticker = sector_tickers) %>%
  left_join(test_daily_signals, by = c("date", "ticker")) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  fill(signal, .direction = "down") %>%
  mutate(signal = replace_na(signal, 0L)) %>%
  ungroup()

test_exec_dates_set <- unique(test_exec_map$exec_date)

test_daily_rets <- raw_data %>%
  filter(ticker %in% sector_tickers, date %in% test_dates) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(
    ret_cc  = Close / dplyr::lag(Close) - 1,
    ret_oc  = Close / Open - 1,
    is_exec = date %in% test_exec_dates_set,
    ret_cc  = if_else(is_exec, ret_oc, ret_cc)
  ) %>%
  ungroup() %>%
  select(date, ticker, ret_cc)

test_rebal_costs <- compute_weights(test_daily_pos) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(prev_weight = dplyr::lag(weight, default = 0)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(tc = sum(abs(weight - prev_weight), na.rm = TRUE) * TC_BPS, .groups = "drop")

test_pnl1 <- compute_weights(test_daily_pos) %>%
  left_join(test_daily_rets, by = c("date", "ticker")) %>%
  group_by(date) %>%
  summarise(ret_gross = sum(weight * replace_na(ret_cc, 0), na.rm = TRUE),
            n_long = first(n_long), .groups = "drop") %>%
  left_join(test_rebal_costs, by = "date") %>%
  mutate(tc = replace_na(tc, 0),
         ret   = ret_gross - tc,
         cumeq = cumprod(1 + ret))

test_monthly_spread <- test_daily_pos %>%
  left_join(test_daily_rets, by = c("date", "ticker")) %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym, signal) %>%
  summarise(avg_ret = mean(ret_cc, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = signal, values_from = avg_ret,
              names_prefix = "sig_") %>%
  arrange(ym) %>%
  mutate(realized_spread = replace_na(sig_1, 0) - replace_na(sig_0, 0)) %>%
  pull(realized_spread)

n_train_months <- length(monthly_spread)
test_kf_mod <- dlmModPoly(1, dV = kf_train$V, dW = kf_train$W)
test_kf_mod$m0 <- kf_train$state[n_train_months]
test_kf_filt <- dlmFilter(test_monthly_spread, test_kf_mod)

test_kf_state <- drop(test_kf_filt$m[-1])
test_kf_innov <- drop(test_monthly_spread - test_kf_filt$f)

test_month_ends <- test_pnl1 %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym) %>% slice_max(date, n = 1) %>% ungroup() %>%
  arrange(date) %>% pull(ym)

test_kf_month_scale <- tibble(
  ym       = test_month_ends,
  kf_state = test_kf_state[seq_along(test_month_ends)],
  kf_innov = test_kf_innov[seq_along(test_month_ends)]
) %>%
  mutate(
    kf_scale = case_when(
      is.na(kf_state)          ~ 1.0,
      kf_state >= KF_HIGH_ABS  ~ 1.0,
      kf_state >= KF_LOW_ABS   ~ 0.5,
      TRUE                     ~ 0.0
    ),
    kf_scale = case_when(
      !is.na(kf_innov) & kf_innov >  abs(KF_HIGH_ABS) ~ pmin(kf_scale + 0.5, 1.0),
      !is.na(kf_innov) & kf_innov < -abs(KF_HIGH_ABS) ~ pmax(kf_scale - 0.5, 0.0),
      TRUE ~ kf_scale
    ),
    next_ym_scale = dplyr::lead(kf_scale, default = 1.0)
  ) %>%
  mutate(ym = floor_date(ym %m+% months(1), "month")) %>%
  select(ym, next_ym_scale)

test_daily_pos_scaled <- test_daily_pos %>%
  mutate(ym = floor_date(date, "month")) %>%
  left_join(test_kf_month_scale, by = "ym") %>%
  mutate(next_ym_scale = replace_na(next_ym_scale, 1.0)) %>%
  select(-ym)

test_rc_scaled <- compute_weights(test_daily_pos_scaled) %>%
  mutate(weight_s = weight * next_ym_scale) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(prev_ws = dplyr::lag(weight_s, default = 0)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(tc = sum(abs(weight_s - prev_ws), na.rm = TRUE) * TC_BPS, .groups = "drop")

test_pnl2 <- compute_weights(test_daily_pos_scaled) %>%
  mutate(weight_s = weight * next_ym_scale) %>%
  left_join(test_daily_rets, by = c("date", "ticker")) %>%
  group_by(date) %>%
  summarise(
    ret_sectors = sum(weight_s * replace_na(ret_cc, 0), na.rm = TRUE),
    kf_scale    = first(next_ym_scale),
    .groups     = "drop"
  ) %>%
  left_join(test_rc_scaled, by = "date") %>%
  left_join(rf_daily,       by = "date") %>%
  arrange(date) %>%
  mutate(
    tc       = replace_na(tc, 0),
    rf_daily = replace_na(rf_daily, 0),
    ret_gross = ret_sectors + (1 - kf_scale) * rf_daily,
    fully_flat = kf_scale == 0,
    rf_tc = if_else(
      (fully_flat & !dplyr::lag(fully_flat, default = FALSE)) |
      (!fully_flat &  dplyr::lag(fully_flat, default = FALSE)),
      TC_BPS, 0),
    ret   = ret_gross - tc - rf_tc,
    cumeq = cumprod(1 + ret)
  )

test_spy_bh <- raw_data %>%
  filter(ticker == "SPY", date %in% test_dates) %>%
  arrange(date) %>%
  mutate(ret_cc    = Close / dplyr::lag(Close) - 1,
         cumeq_spy = cumprod(1 + replace_na(ret_cc, 0))) %>%
  select(date, cumeq_spy, ret_cc_spy = ret_cc)

test_bench_def  <- ew_benchmark(defensive_tickers, test_dates, raw_data)
test_bench_cyc  <- ew_benchmark(cyclical_tickers,  test_dates, raw_data)
test_bench_5050 <- spy_top3_bench(test_dates, test_rebal_dates, raw_data,
                                   sector_ret, sector_tickers)

test_perf <- test_pnl2 %>%
  select(date, cumeq, ret, rf_daily, kf_scale) %>%
  mutate(kf_scale = as.numeric(kf_scale)) %>%
  left_join(test_spy_bh, by = "date") %>%
  left_join(test_bench_def  %>% select(date, cumeq_def  = cumeq), by = "date") %>%
  left_join(test_bench_cyc  %>% select(date, cumeq_cyc  = cumeq), by = "date") %>%
  left_join(test_bench_5050 %>% select(date, cumeq_5050 = cumeq), by = "date")
```

```{r test-stats-compute}
n_test_days  <- nrow(test_perf)
n_test_years <- n_test_days / 252

test_rets  <- replace_na(test_perf$ret, 0)
test_rf    <- replace_na(test_perf$rf_daily, 0)
test_spy   <- replace_na(test_perf$ret_cc_spy, 0)
excess_t   <- test_rets - test_rf
excess_spy_test <- test_spy  - test_rf

test_peak     <- cummax(test_perf$cumeq)
test_peak_spy <- cummax(test_perf$cumeq_spy)
test_mdd      <- min((test_perf$cumeq     - test_peak)     / test_peak,     na.rm = TRUE)
test_mdd_spy  <- min((test_perf$cumeq_spy - test_peak_spy) / test_peak_spy, na.rm = TRUE)
test_total    <- last(test_perf$cumeq)     - 1
test_total_spy<- last(test_perf$cumeq_spy) - 1
test_sharpe   <- (mean(excess_t, na.rm=TRUE)/sd(excess_t, na.rm=TRUE))*sqrt(252)
test_sharpe_spy <- (mean(excess_spy_test, na.rm=TRUE)/sd(excess_spy_test, na.rm=TRUE))*sqrt(252)
test_omega    <- omega_fn(test_rets)
test_omega_spy <- omega_fn(test_spy)
```

<!----------------------------------------------------------------------
  ===================================================================
   PRESENTATION BEGINS HERE — all computation is complete above
  ===================================================================
----------------------------------------------------------------------->

# Insights

- A long-only sector rotation strategy with a two-layer clustering architecture combined with a Kalman filter for dynamic exposure scaling. 

- The first layer clusters Kalman-smoothed macro indicators for regime detection, while the second layer clusters for the best-performing sectors within the detected regime. A Kalman filter compared model predictions with realized returns to scale portfolio exposure accordingly.

- Over the training period, this strategy achieved a total return of `r scales::percent(last(training_perf$cumeq_kf) - 1, accuracy = 0.1)` with max drawdown of `r scales::percent(max_dd_kf, accuracy = 0.1)` compared to the SPY benchmark at `r scales::percent(total_spy, accuracy = 0.1)` and max drawdown `r scales::percent(max_dd_spy, accuracy = 0.1)`. The testing period yielded `r scales::percent(test_total, accuracy = 0.1)` vs SPY at `r scales::percent(test_total_spy, accuracy = 0.1)`.

---

# Problem Breakdown

To maximize gains from sector rotation, understanding the dominant regime and its subsequent sector behavior is important:

```{mermaid}
flowchart TD
    C[Macro Data via API<br/>VIX, VRP, Yield Curve, Breakeven] --> E[Layer 1: Regime Detection<br/>K-means K=3 on 4 KF-smoothed macro features<br/>Expanding window]
    E -->|Regime detection| J[Layer 2: Sector Clustering<br/>K-means K=2 on<br/>regime_ret, regime_vol,<br/>regime_sharpe, beta_252]
    J --> K[Best cluster — highest avg return<br/>Top 3 by momentum_21 → signal = 1]
    K --> O[KF Predict-Correct — kf_exposure<br/>State = prediction quality<br/>Innovation = surprise]
    O --> P{KF State Level}
    P -->|α ≥ KF_HIGH| Q[Full exposure 100%]
    P -->|KF_LOW ≤ α < KF_HIGH| R[Partial exposure 50%<br/>50% earns rf]
    P -->|α < KF_LOW| S[Flat 0%<br/>100% earns rf]
    V[VIX > 25?] -->|Yes| G[All signals = 0<br/>Fully flat]
    Q --> T[Portfolio P&L<br/>cumeq = cumprod 1+ret]
    R --> T
    S --> T
    G --> T
```

Key design decisions:

- **Two-layer separation:** The same sector can outperform in one regime and underperform in another. Regime-conditioning the sector clustering captures this.
- **KF predict-correct replaces ad-hoc stops:** The Kalman filter tracks model prediction quality directly, scaling down exposure when the clustering model stops working before large losses accumulate.

---

# Model Construction

## Data & Assumptions

```{r universe-table}
tibble(
  Ticker = c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
             "XLK", "XLP", "XLRE", "XLU", "XLV", "XLY"),
  Sector = c("S&P 500 (Benchmark)", "Materials", "Communication Services",
             "Energy", "Financials", "Industrials", "Technology",
             "Consumer Staples", "Real Estate", "Utilities",
             "Health Care", "Consumer Discretionary"),
  Group = c("Benchmark", "Cyclical", "Cyclical", "Cyclical", "Cyclical",
            "Cyclical", "Cyclical", "Defensive", "Defensive", "Defensive",
            "Defensive", "Cyclical"),
  `Available From` = c("1993", "1998", "2018", "1998", "1998", "1998",
                        "1998", "1998", "2015", "1998", "1998", "1998")
) %>%
  gt() %>%
  tab_header(title = "Asset List") %>%
  tab_style(
    style = list(cell_fill(color = "#f8f9fa"), cell_text(weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Ticker, rows = Ticker == "SPY")
  ) %>%
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(rows = `Available From` %in% c("2015", "2018"))
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    heading.subtitle.font.size = px(13),
    table.border.top.color = "#4e79a7",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "#4e79a7"
  )
```

We pair this data with the following assumptions:

- Transaction costs are on average 5 basis points per side.
- Rebalancing takes in close values at the end of the month and execute at open for the beginning of next month. 
- Clustering refits on prior data at each rebalancing.

## Layer 1 -- Regime Detection

Layer 1 clusters trading dates into macro regimes using K-means on 4 Kalman-smoothed indicators. Each indicator is smoothed via estimated noise parameters.

```{r layer1-features-table}
tibble(
  Feature = c("VIX", "VRP (21-day rolling)", "Yield Curve", "Breakeven Inflation"),
  Source = c("Yahoo", "VIX - SPY", "TNX - DGS2 (10Y-2Y)", "T10YIE (FRED)"),
  `Economic Meaning` = c(
    "Market-wide implied fear / expected volatility",
    "Fear premium vs realized stress -- positive = overpriced risk",
    "Monetary policy stance / recession signal (negative = inverted)",
    "Market-implied 10Y avg inflation -- orthogonal regime dimension"
  )
) %>%
  gt() %>%
  tab_header(title = "Layer 1 Clustering Features") %>%
  tab_style(
    style = list(cell_fill(color = "#f8f9fa"), cell_text(weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    heading.subtitle.font.size = px(13),
    table.border.top.color = "#4e79a7",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "#4e79a7"
  )
```

K = 3 is selected for Layer 1 based on the elbow plot inflection and domain reasoning (3 regimes: risk-on, risk-off, transition). 

```{r layer1-pca}
pca_res  <- prcomp(cluster_mat, center = FALSE, scale. = FALSE)
pca_vals <- as_tibble(pca_res$x[, 1:2]) %>%
  bind_cols(cluster_assignments %>% select(date, km_cluster)) %>%
  mutate(km_cluster = as.factor(km_cluster))

pct_var <- round(summary(pca_res)$importance[2, 1:2] * 100, 1)

# Build axis labels from top loadings
loadings <- pca_res$rotation[, 1:2]
pc1_order <- sort(abs(loadings[, 1]), decreasing = TRUE)
pc2_order <- sort(abs(loadings[, 2]), decreasing = TRUE)
pc1_label <- paste0("PC1 (", pct_var[1], "%) \u2014 ",
                    paste(names(pc1_order)[1:2], collapse = ", "))
pc2_label <- paste0("PC2 (", pct_var[2], "%) \u2014 ",
                    paste(names(pc2_order)[1:2], collapse = ", "))

p_pca <- ggplot(pca_vals, aes(x = PC1, y = PC2, colour = km_cluster)) +
  geom_point(alpha = 0.4, size = 1.0) +
  scale_colour_manual(values = c("1" = "#4e79a7", "2" = "#f28e2b", "3" = "#59a14f")) +
  labs(
    title = "Layer 1 Regime Clusters -- PCA Biplot",
    subtitle = paste0("PC1 ", pct_var[1], "% | PC2 ", pct_var[2], "% of variance"),
    x = pc1_label,
    y = pc2_label,
    colour = "Regime"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

ggplotly(p_pca, tooltip = c("x", "y", "colour"))
```

```{r layer1-profiles}
profile_data <- cluster_mat %>%
  as_tibble() %>%
  mutate(km_cluster = factor(cluster_assignments$km_cluster)) %>%
  pivot_longer(-km_cluster, names_to = "feature", values_to = "value") %>%
  group_by(km_cluster, feature) %>%
  summarise(mean_z = mean(value, na.rm = TRUE), .groups = "drop")

ggplot(profile_data, aes(x = km_cluster, y = feature, fill = mean_z)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = round(mean_z, 2)), size = 3) +
  scale_fill_gradient2(low = "#4e79a7", mid = "white", high = "#e15759",
                       midpoint = 0, name = "Mean Z") +
  labs(title = "Regime Profiles -- Mean Macro Feature Values",
       subtitle = "Red = above average | Blue = below average",
       x = "Regime Cluster", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), axis.text.y = element_text(size = 9))
```

Based on the above graphs:

- Regime 1 is a normal, risk-on period.

- Regime 2 experiences low volatility expansion.

- Regime 3 indicates crisis and economic stress.



## Layer 2 -- Sector Clustering

Within the detected regime, Layer 2 clusters the 11 sector ETFs into K=2 groups using regime-conditional features. The cluster with the highest average return is selected, and the top 3 sectors by 21-day momentum within that cluster are signalled long.

```{r layer2-features-table}
tibble(
  Feature = c("regime_ret", "regime_vol", "regime_sharpe", "beta_252"),
  Computation = c(
    "Mean daily return over regime dates",
    "SD of daily returns over regime dates",
    "regime_ret / regime_vol",
    "Current day's 252d rolling beta to SPY"
  ),
  Purpose = c(
    "Sector performance in this specific macro environment",
    "Sector risk in this environment",
    "Risk-adjusted performance within the regime",
    "Cyclical/defensive characterization (not regime-averaged)"
  )
) %>%
  gt() %>%
  tab_header(title = "Layer 2 Clustering Features", subtitle = "Computed per-sector within the current regime") %>%
  tab_style(
    style = list(cell_fill(color = "#f8f9fa"), cell_text(weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    heading.subtitle.font.size = px(13),
    table.border.top.color = "#4e79a7",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "#4e79a7"
  )
```

K = 2 is selected for Layer 2 to cleanly separate outperforming vs underperforming sectors within each regime.

```{r layer2-silhouette-bar}
#| fig-height: 5
ggplot(sil_avg, aes(x = reorder(ticker, -order), y = sil_width, fill = cluster)) +
  geom_col(width = 0.8) +
  geom_hline(yintercept = avg_sil, linetype = "dashed", colour = "grey50", linewidth = 0.6) +
  annotate("text", x = nrow(sil_avg), y = avg_sil + 0.02,
           label = sprintf("Avg = %.2f", avg_sil),
           hjust = 1, size = 3.5, colour = "grey50") +
  scale_fill_manual(values = c("1" = "#e15759", "2" = "#59a14f"),
                    labels = c("1" = "Cluster 1", "2" = "Cluster 2"),
                    name = "Sector Cluster") +
  coord_flip() +
  labs(title = "Per-Sector Silhouette Width (K = 2)",
       subtitle = "Averaged across 20 sampled rebalancing dates (training period)",
       x = NULL, y = "Silhouette Width") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "right")
```


## Kalman Filter & Trade Signal Methodology

The Kalman filter serves two distinct roles. First, `kf_level()` smooths the macro indicators before Layer 1 clustering. Second, `kf_exposure()` implements a predict-correct framework that tracks whether the clustering model's predictions are in line with incoming data.

```{r kf-params-table}
tibble(
  Parameter = c("Model", "Observation (y_t)", "State (alpha_t)", "V, W"),
  Value = c(
    "Local level (random walk + noise), dlmModPoly(1)",
    "Monthly realized return spread: mean(ret[signal=1]) - mean(ret[signal=0])",
    "Latent prediction quality -- how well clustering picks outperform",
    "MLE on training data, frozen for test period"
  )
) %>%
  gt() %>%
  tab_header(title = "KF Predict-Correct Parameters") %>%
  tab_style(
    style = list(cell_fill(color = "#f8f9fa"), cell_text(weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    table.border.top.color = "#4e79a7",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "#4e79a7"
  )
```

```{r kf-exposure-table}
tibble(
  `KF State` = c("alpha >= KF_HIGH", "KF_LOW <= alpha < KF_HIGH", "alpha < KF_LOW"),
  Exposure = c("100% (full)", "50% (partial)", "0% (flat)"),
  Action = c(
    "Predictions confirmed -- invest fully in selected sectors",
    "Uncertain -- half in sector picks, half earns risk-free rate",
    "Predictions failing -- all capital earns risk-free rate (IRX / 252)"
  )
) %>%
  gt() %>%
  tab_header(title = "Exposure Mapping", subtitle = "KF state determines portfolio allocation") %>%
  tab_style(
    style = list(cell_fill(color = "#f8f9fa"), cell_text(weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_fill(color = "#d4edda"),
    locations = cells_body(rows = 1)
  ) %>%
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(rows = 2)
  ) %>%
  tab_style(
    style = cell_fill(color = "#f8d7da"),
    locations = cells_body(rows = 3)
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    heading.subtitle.font.size = px(13),
    table.border.top.color = "#4e79a7",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "#4e79a7"
  )
```

Large positive innovation (market outperformed predictions) increases exposure while large negative innovation (model failing) decreases exposure. The derived Kalman filter's high and low thresholds also affects the difference between mean daily returns of long vs. flat sectors from clustering as follows:

```{r kf-diagnostics}
#| fig-height: 6
kf_diag <- tibble(
  ym        = month_ends,
  state     = kf_train$state[seq_along(month_ends)],
  innov     = kf_train$innovation[seq_along(month_ends)],
  realized  = monthly_spread[seq_along(month_ends)]
) %>% filter(!is.na(state))

fig_state <- plot_ly(kf_diag, x = ~ym) %>%
  add_lines(y = ~state, name = "KF state",
            line = list(color = "#4e79a7", width = 2)) %>%
  add_segments(x = ~min(ym), xend = ~max(ym),
               y = KF_HIGH_ABS, yend = KF_HIGH_ABS, inherit = FALSE,
               line = list(color = "#59a14f", width = 1, dash = "dot"),
               name = paste0("KF_HIGH (q=", KF_HIGH_Q, ")")) %>%
  add_segments(x = ~min(ym), xend = ~max(ym),
               y = KF_LOW_ABS, yend = KF_LOW_ABS, inherit = FALSE,
               line = list(color = "#e15759", width = 1, dash = "dot"),
               name = paste0("KF_LOW (q=", KF_LOW_Q, ")")) %>%
  layout(
    title = list(text = "KF State - Clustering Prediction Quality Over Time", font = list(size = 14)),
    xaxis = list(title = "Time", type = "date"),
    yaxis = list(title = "Daily Return"),
    legend = list(x = 0.01, y = 0.99, bgcolor = "rgba(255,255,255,0.7)")
  )

fig_state
```

---

# Training & Optimization

## Training Setup

```{r fixed-params-table}
tibble(
  Parameter = c("K_CLUSTERS", "K_SECTORS", "N_LONG",
                "VIX_THRESHOLD", "MAX_REGIME_DAYS", "TC_BPS"),
  Value = c("3", "2", "3", "25", "120", "0.0005 (5 bps/side)"),
  Rationale = c(
    "Silhouette-optimal for Layer 1",
    "Silhouette-optimal size for Layer 2",
    "Max sectors long from best cluster",
    "Hard gate -- fully flat when VIX exceeds",
    "Cap on regime history for responsiveness (~6 months)",
    "Transaction cost per side on weight turnover"
  )
) %>%
  gt() %>%
  tab_header(title = "Fixed Parameters") %>%
  tab_style(
    style = list(cell_fill(color = "#f8f9fa"), cell_text(weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    table.border.top.color = "#4e79a7",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "#4e79a7"
  )
```


## Training Performance vs Benchmarks

```{r training-cumeq-chart}
fig_cumeq <- plot_ly(training_perf, x = ~date) %>%
  add_lines(y = ~cumeq_kf, name = "Strategy (Clustering + KF)",
            line = list(color = "#59a14f", width = 2)) %>%
  add_lines(y = ~cumeq_spy, name = "SPY Buy & Hold",
            line = list(color = "#e15759", width = 1.5)) %>%
  add_lines(y = ~cumeq_5050, name = "50% SPY + 50% Top 3 Sectors",
            line = list(color = "#4e79a7", width = 1)) %>%
  add_lines(y = ~cumeq_def, name = "100% Defensive (XLU/XLP/XLV/XLRE)",
            line = list(color = "#f28e2b", width = 1)) %>%
  add_lines(y = ~cumeq_cyc, name = "100% Cyclical (XLK/XLY/XLF/XLI/XLB/XLE/XLC)",
            line = list(color = "#b07aa1", width = 1)) %>%
  layout(
    title = list(text = "Training Period Cumulative Return -- Strategy vs Benchmarks",
                 font = list(size = 14)),
    xaxis = list(title = NULL, type = "date"),
    yaxis = list(title = "Cumulative Return (1.0 = start)", tickformat = ".2f"),
    legend = list(x = 0.01, y = 0.99, bgcolor = "rgba(255,255,255,0.7)"),
    hovermode = "x unified"
  )

fig_cumeq
```



## Optimization & Risk Appetite

We can vary the Kalman filter's low and high thresholds to optimize our strategy.


```{r opt-heatmap}
heatmap_data <- opt_results %>%
  mutate(KF_HIGH_label = sprintf("%.0f%%", KF_HIGH_Q * 100),
         KF_LOW_label  = sprintf("%.0f%%", KF_LOW_Q  * 100))

p_heat <- ggplot(heatmap_data,
                 aes(x = KF_HIGH_label, y = KF_LOW_label, fill = omega)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = round(omega, 2)), size = 2.5) +
  scale_fill_gradient(low = "white", high = "#59a14f", name = "Omega") +
  labs(
    title = paste0("Omega by KF Thresholds (K=", K_CLUSTERS, ", N_LONG=", N_LONG, ", VIX<=", VIX_THRESHOLD, ")"),
    x = "KF High Quantile", y = "KF Low Quantile"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

p_heat
```

```{r opt-results-gt}
opt_results %>%
  arrange(desc(omega)) %>%
  head(10) %>%
  gt() %>%
  tab_header(
    title = "Top 10 KF Threshold Combinations by Omega",
    subtitle = paste0("Training period | K=", K_CLUSTERS, ", N_LONG=", N_LONG, ", VIX<=", VIX_THRESHOLD, " (fixed)")
  ) %>%
  fmt_percent(columns = c(KF_HIGH_Q, KF_LOW_Q, total_ret, cagr, max_dd), decimals = 1) %>%
  fmt_number(columns = c(sharpe, omega), decimals = 3) %>%
  cols_label(
    KF_HIGH_Q = "KF High Q", KF_LOW_Q = "KF Low Q",
    total_ret = "Total Ret", cagr = "CAGR",
    sharpe = "Sharpe", omega = "Omega", max_dd = "Max DD"
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#f8f9fa"), cell_text(weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_fill(color = "#d4edda"),
    locations = cells_body(rows = 1)
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    heading.subtitle.font.size = px(13),
    table.border.top.color = "#4e79a7",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "#4e79a7"
  )
```

We optimize for Omega in order to ensure that sustainable returns take into account tail events. Based on this criteria, the highlighted above will be used for testing.

---

# Backtesting

Levels reset at the beginning of backtesting using optimized settings:

```{r test-cumeq-chart}
plot_ly(test_perf, x = ~date) %>%
  add_lines(y = ~cumeq, name = "Strategy (Clustering + KF)",
            line = list(color = "#59a14f", width = 2)) %>%
  add_lines(y = ~cumeq_spy, name = "SPY Buy & Hold",
            line = list(color = "#e15759", width = 1.5)) %>%
  add_lines(y = ~cumeq_5050, name = "50% SPY + 50% Top 3 Sectors",
            line = list(color = "#4e79a7", width = 1)) %>%
  add_lines(y = ~cumeq_def, name = "100% Defensive (XLU/XLP/XLV/XLRE)",
            line = list(color = "#f28e2b", width = 1)) %>%
  add_lines(y = ~cumeq_cyc, name = "100% Cyclical (XLK/XLY/XLF/XLI/XLB/XLE/XLC)",
            line = list(color = "#b07aa1", width = 1)) %>%
  layout(
    title = list(text = "Test Period Cumulative Return (Out-of-Sample)",
                 font = list(size = 14)),
    xaxis = list(title = NULL, type = "date"),
    yaxis = list(title = "Cumulative Return (1.0 = start)", tickformat = ".2f"),
    hovermode = "x unified",
    legend = list(x = 0.01, y = 0.99, bgcolor = "rgba(255,255,255,0.8)")
  )
```

```{r test-stats-gt}
tibble(
  Metric = c("Total Return", "CAGR", "Ann. Volatility", "Sharpe Ratio", "Omega (L=0)",
             "Max Drawdown", "MDD Date", "Period"),
  Strategy = c(
    scales::percent(test_total, accuracy = 0.1),
    scales::percent((1 + test_total)^(1/n_test_years) - 1, accuracy = 0.1),
    scales::percent(sd(test_rets, na.rm = TRUE) * sqrt(252), accuracy = 0.1),
    round(test_sharpe, 3),
    round(test_omega, 3),
    scales::percent(test_mdd, accuracy = 0.1),
    format(test_perf$date[which.min((test_perf$cumeq - test_peak)/test_peak)], "%Y-%m-%d"),
    paste(format(TEST_START, "%Y-%m-%d"), "--", format(TEST_END, "%Y-%m-%d"))
  ),
  `SPY B&H` = c(
    scales::percent(test_total_spy, accuracy = 0.1),
    scales::percent((1 + test_total_spy)^(1/n_test_years) - 1, accuracy = 0.1),
    scales::percent(sd(test_spy, na.rm = TRUE) * sqrt(252), accuracy = 0.1),
    round(test_sharpe_spy, 3),
    round(test_omega_spy, 3),
    scales::percent(test_mdd_spy, accuracy = 0.1),
    format(test_perf$date[which.min((test_perf$cumeq_spy - test_peak_spy)/test_peak_spy)], "%Y-%m-%d"),
    paste(format(TEST_START, "%Y-%m-%d"), "--", format(TEST_END, "%Y-%m-%d"))
  )
) %>%
  gt() %>%
  tab_header(
    title = "Test Period Performance"
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#f8f9fa"), cell_text(weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Metric)
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    heading.subtitle.font.size = px(13),
    table.border.top.color = "#4e79a7",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "#4e79a7"
  )
```

---

# Improvements

Based on the above results, potential enhancements can be explored to improve this strategy:

- Consider different variables for clustering each layer so the model inherently chooses long sectors that outperform flat sectors.
- Introduce simulations in backtesting to improve robustness.
- Using ensemble methods and model averaging reduces model risk and illustrates more scenarios under this decision ruleset.

---

# Appendix

## Historical Price Levels

```{r appendix-price-levels}
display_data <- raw_data %>%
  filter(date >= TRAIN_START) %>%
  group_by(ticker) %>%
  arrange(date) %>%
  mutate(level = 100 * Close / first(Close)) %>%
  ungroup() %>%
  select(date, ticker, level) %>%
  pivot_wider(names_from = ticker, values_from = level)

p <- display_data %>%
  pivot_longer(-date, names_to = "ticker", values_to = "level") %>%
  mutate(ticker = factor(ticker, levels = tickers)) %>%
  ggplot(aes(x = date, y = level)) +
  annotate("rect", xmin = TEST_START, xmax = TEST_END,
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +
  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.4) +
  geom_line(colour = "steelblue", linewidth = 0.4) +
  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +
  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  labs(title = "SPDR Sector ETF Historical Levels (Indexed to 100)",
       x = NULL, y = "Indexed Level") +
  theme_minimal(base_size = 10) +
  theme(strip.text = element_text(size = 7.5, face = "bold"),
        axis.text.x = element_text(size = 7, angle = 30, hjust = 1),
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))

ggplotly(p, tooltip = c("x", "y")) %>% layout(showlegend = FALSE)
```

## Treasury Yields & Breakeven Inflation

```{r appendix-rates}
rates_long <- rates %>%
  pivot_longer(-date, names_to = "series", values_to = "value") %>%
  mutate(series = recode(series,
    y10        = "10Y Treasury",
    y2         = "2Y Treasury",
    yield_curve = "Yield Curve (10Y - 2Y)",
    breakeven  = "Breakeven Inflation (T10YIE)"))

p_rates <- ggplot(rates_long, aes(x = date, y = value, colour = series)) +
  annotate("rect", xmin = TEST_START,
           xmax = max(rates_long$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +
  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.4) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.3) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  scale_colour_manual(values = c(
    "10Y Treasury" = "#4e79a7", "2Y Treasury" = "#76b7b2",
    "Yield Curve (10Y - 2Y)" = "#e15759",
    "Breakeven Inflation (T10YIE)" = "#f28e2b")) +
  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +
  labs(title = "Treasury Yields, Yield Curve & Breakeven Inflation",
       x = NULL, y = "Rate (%)", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))

ggplotly(p_rates, tooltip = c("x", "y", "colour")) %>%
  layout(legend = list(orientation = "h", y = 1.08))
```

## SPY Rolling Vol vs VIX

```{r appendix-spy-vol}
spy_vol <- features_wide %>%
  select(date, `21d Realised` = SPY_rvol_21, VIX = vix, VRP = vrp) %>%
  pivot_longer(-date, names_to = "series", values_to = "vol") %>%
  mutate(series = factor(series, levels = c("VIX", "21d Realised", "VRP")))

p_spy <- ggplot(spy_vol, aes(x = date, y = vol, colour = series)) +
  annotate("rect", xmin = TEST_START, xmax = max(spy_vol$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +
  geom_vline(xintercept = TEST_START, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60", linewidth = 0.3) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  scale_colour_manual(values = c("VIX" = "#e15759", "21d Realised" = "#4e79a7", "VRP" = "#f28e2b")) +
  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +
  labs(title = "SPY: VIX vs 21d Realised Volatility & VRP",
       subtitle = "VRP = VIX - 21d Realised Vol | Positive = fear premium",
       x = NULL, y = "Annualised Volatility (%)", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))

ggplotly(p_spy, tooltip = c("x", "y", "colour")) %>%
  layout(legend = list(orientation = "h", y = 1.08))
```

## Sector Vol Spread Z-Score

```{r appendix-sector-vol}
sector_vol <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  mutate(ticker = factor(ticker, levels = sector_tickers))

p_sec <- ggplot(sector_vol, aes(x = date, y = spread_z)) +
  annotate("rect", xmin = TEST_START, xmax = max(sector_vol$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +
  geom_vline(xintercept = TEST_START, linetype = "dashed", colour = "grey40", linewidth = 0.3) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.3) +
  geom_line(colour = "#4e79a7", linewidth = 0.35, alpha = 0.85) +
  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +
  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  labs(title = "Sector Vol Spread Z-Score (sector rvol - VIX, normalised)",
       x = NULL, y = "Z-Score") +
  theme_minimal(base_size = 10) +
  theme(strip.text = element_text(size = 8, face = "bold"),
        axis.text.x = element_text(size = 7, angle = 30, hjust = 1),
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))

ggplotly(p_sec, tooltip = c("x", "y")) %>% layout(showlegend = FALSE)
```

## Rolling Beta to SPY

```{r appendix-beta}
beta_data <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  mutate(ticker = factor(ticker, levels = sector_tickers))

p_beta <- ggplot(beta_data, aes(x = date, y = beta_252)) +
  annotate("rect", xmin = TEST_START, xmax = max(beta_data$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +
  geom_vline(xintercept = TEST_START, linetype = "dashed", colour = "grey40", linewidth = 0.3) +
  geom_hline(yintercept = 1, linetype = "dotted", colour = "grey50", linewidth = 0.3) +
  geom_line(colour = "#59a14f", linewidth = 0.35, alpha = 0.85) +
  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +
  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  labs(title = "252d Rolling Beta to SPY", x = NULL, y = "Beta") +
  theme_minimal(base_size = 10) +
  theme(strip.text = element_text(size = 8, face = "bold"),
        axis.text.x = element_text(size = 7, angle = 30, hjust = 1),
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))

ggplotly(p_beta, tooltip = c("x", "y")) %>% layout(showlegend = FALSE)
```

## 21d Rolling Log Return

```{r appendix-logret}
logret_data <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  mutate(ticker = factor(ticker, levels = sector_tickers))

p_logret <- ggplot(logret_data, aes(x = date, y = log_ret_21)) +
  annotate("rect", xmin = TEST_START, xmax = max(logret_data$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +
  geom_vline(xintercept = TEST_START, linetype = "dashed", colour = "grey40", linewidth = 0.3) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.3) +
  geom_line(colour = "#b07aa1", linewidth = 0.35, alpha = 0.85) +
  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +
  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(title = "21d Rolling Log Return (Sector Momentum)", x = NULL, y = "Log Return") +
  theme_minimal(base_size = 10) +
  theme(strip.text = element_text(size = 8, face = "bold"),
        axis.text.x = element_text(size = 7, angle = 30, hjust = 1),
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))

ggplotly(p_logret, tooltip = c("x", "y")) %>% layout(showlegend = FALSE)
```

## Regime-Conditional Sector Profiles

```{r appendix-sector-profiles}
#| fig-height: 8
regime_sector_profiles <- signals_raw %>%
  filter(!is.na(regime), !is.na(avg_ret)) %>%
  left_join(
    sector_ret %>% select(date, ticker, beta_252),
    by = c("rebal_date" = "date", "ticker")
  ) %>%
  group_by(regime, ticker) %>%
  summarise(avg_ret = mean(avg_ret, na.rm = TRUE),
            avg_beta = mean(beta_252, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(regime_label = paste0("Regime ", regime),
         ticker = factor(ticker, levels = sector_tickers))

p_ret <- ggplot(regime_sector_profiles,
                aes(x = reorder(ticker, avg_ret), y = avg_ret, fill = regime_label)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("Regime 1" = "#4e79a7", "Regime 2" = "#f28e2b",
                                "Regime 3" = "#59a14f"), name = NULL) +
  facet_wrap(~regime_label, ncol = 1, scales = "free_x") +
  coord_flip() +
  labs(title = "Average Regime Return by Sector", x = NULL, y = "Mean Regime Return") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold"), legend.position = "none",
        strip.text = element_text(face = "bold"))

p_beta_app <- ggplot(regime_sector_profiles,
                 aes(x = reorder(ticker, avg_beta), y = avg_beta, fill = regime_label)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("Regime 1" = "#4e79a7", "Regime 2" = "#f28e2b",
                                "Regime 3" = "#59a14f"), name = NULL) +
  facet_wrap(~regime_label, ncol = 1, scales = "free_x") +
  coord_flip() +
  labs(title = "Average Beta by Sector", x = NULL, y = "Mean Beta (252d)") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold"), legend.position = "none",
        strip.text = element_text(face = "bold"))

p_ret + p_beta_app
```

## KF Exposure Decisions by Regime

```{r appendix-kf-exposure-regime}
#| fig-height: 5
kf_regime <- daily_pos_kf %>%
  mutate(ym = floor_date(date, "month")) %>%
  distinct(ym, next_ym_scale) %>%
  left_join(
    signals_raw %>%
      distinct(rebal_date, regime) %>%
      mutate(ym = floor_date(rebal_date, "month")),
    by = "ym"
  ) %>%
  filter(!is.na(regime), !is.na(next_ym_scale)) %>%
  mutate(
    regime_label = paste0("Regime ", regime),
    exposure = case_when(
      near(next_ym_scale, 0)   ~ "0% (flat)",
      near(next_ym_scale, 0.5) ~ "50% (partial)",
      TRUE                     ~ "100% (full)"
    ),
    exposure = factor(exposure, levels = c("0% (flat)", "50% (partial)", "100% (full)"))
  )

kf_regime_summary <- kf_regime %>%
  count(regime_label, exposure) %>%
  group_by(regime_label) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

ggplot(kf_regime_summary, aes(x = exposure, y = pct, fill = exposure)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("0% (flat)" = "#e15759", "50% (partial)" = "#f28e2b",
                                "100% (full)" = "#59a14f"), name = NULL) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~regime_label, ncol = 3) +
  labs(title = "KF Exposure Distribution by Regime", x = NULL, y = "% of Months") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "none",
        strip.text = element_text(face = "bold"))
```

## Monthly Return Distribution

```{r appendix-monthly-dist}
monthly_perf <- training_perf %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym) %>%
  summarise(
    strat_monthly = prod(1 + portfolio_ret) - 1,
    spy_monthly   = prod(1 + replace_na(ret_cc_spy, 0)) - 1,
    .groups = "drop"
  ) %>%
  pivot_longer(-ym, names_to = "series", values_to = "monthly_ret") %>%
  mutate(series = recode(series,
    strat_monthly = "Strategy",
    spy_monthly   = "SPY Buy & Hold"
  ))

p_monthly_dist <- ggplot(monthly_perf,
                         aes(x = monthly_ret, fill = series, colour = series)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, alpha = 0.4,
                 position = "identity") +
  geom_density(linewidth = 0.8, alpha = 0) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("Strategy" = "#4e79a7", "SPY Buy & Hold" = "#e15759")) +
  scale_colour_manual(values = c("Strategy" = "#4e79a7", "SPY Buy & Hold" = "#e15759")) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(title = "Monthly Return Distribution -- Training Period",
       x = "Monthly Return", y = "Density", fill = NULL, colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")

ggplotly(p_monthly_dist, tooltip = c("x", "density"))
```

## Layer 1 Elbow & Silhouette {#layer-1-elbow--silhouette}

```{r appendix-layer1-diagnostics}
p_elbow <- ggplot(k_metrics, aes(x = k, y = wss)) +
  geom_line(colour = "#4e79a7", linewidth = 0.8) +
  geom_point(colour = "#4e79a7", size = 3) +
  scale_x_continuous(breaks = 2:6) +
  labs(title = "Layer 1 Elbow Plot", x = "K", y = "Total Within-Cluster SS") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

p_sil <- ggplot(k_metrics, aes(x = k, y = silhouette)) +
  geom_line(colour = "#e15759", linewidth = 0.8) +
  geom_point(colour = "#e15759", size = 3) +
  geom_vline(xintercept = 3, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 3.15, y = min(k_metrics$silhouette),
           label = "K = 3 (chosen)", hjust = 0, size = 3.5, colour = "grey40") +
  scale_x_continuous(breaks = 2:6) +
  labs(title = "Layer 1 Average Silhouette Width", x = "K", y = "Mean Silhouette") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

p_elbow + p_sil
```

## KF Innovation (Prediction Errors)

```{r appendix-kf-innovation}
fig_innov <- plot_ly(kf_diag, x = ~ym) %>%
  add_bars(y = ~innov, name = "Innovation",
           marker = list(
             color = if_else(kf_diag$innov >= 0, "#59a14f", "#e15759"),
             opacity = 0.7
           )) %>%
  layout(
    title = list(text = "KF Innovation -- Prediction Errors", font = list(size = 14)),
    xaxis = list(title = NULL, type = "date"),
    yaxis = list(title = "Innovation (obs - predicted)"),
    showlegend = FALSE
  )

fig_innov
```

## Predicted vs Realized Spread (KF Scatter)

```{r appendix-kf-scatter}
pred_monthly <- signals_raw %>%
  select(rebal_date, pred_spread) %>%
  distinct(rebal_date, .keep_all = TRUE) %>%
  mutate(ym = floor_date(rebal_date, "month")) %>%
  select(ym, pred_spread)

scatter_data <- kf_diag %>%
  left_join(pred_monthly, by = "ym") %>%
  filter(!is.na(pred_spread), !is.na(realized))

ref_range <- range(scatter_data$pred_spread, na.rm = TRUE)
ref_line  <- data.frame(x = ref_range, y = ref_range)

fig_scatter <- plot_ly(scatter_data, x = ~pred_spread, y = ~realized,
                       type = "scatter", mode = "markers",
                       marker = list(color = "#4e79a7", size = 8, opacity = 0.7),
                       text = ~format(ym, "%Y-%m"),
                       hovertemplate = "%{text}<br>Pred: %{x:.4f}<br>Real: %{y:.4f}") %>%
  add_lines(data = ref_line, x = ~x, y = ~y, inherit = FALSE,
            line = list(color = "grey50", width = 1, dash = "dash"),
            name = "45 deg line", showlegend = FALSE) %>%
  layout(
    title = list(text = "Predicted vs Realized Return Spread", font = list(size = 14)),
    xaxis = list(title = "Predicted Spread (clustering)"),
    yaxis = list(title = "Realized Spread (market)"),
    showlegend = FALSE
  )

fig_scatter
```

## Regime Timeline (Training)

```{r appendix-regime-timeline}
#| fig-height: 3
regime_timeline <- signals_raw %>%
  distinct(rebal_date, regime) %>%
  mutate(regime = factor(regime))

ggplot(regime_timeline, aes(x = rebal_date, y = 1, fill = regime)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = c("1" = "#4e79a7", "2" = "#f28e2b", "3" = "#59a14f"),
    na.value = "grey85", name = "Regime"
  ) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Layer 1: Market Regime Over Time",
       subtitle = "Macro clustering (VIX, VRP, yield curve, breakeven)",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid = element_blank(), legend.position = "top")
```

## Signal Heatmap (Training)

```{r appendix-signal-heatmap}
signal_heatmap <- signals_raw %>%
  mutate(ticker  = factor(ticker, levels = rev(sector_tickers)),
         signal  = factor(signal, levels = c(-1L, 0L, 1L)))

p_signals <- ggplot(signal_heatmap,
                    aes(x = rebal_date, y = ticker, fill = signal)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = c("-1" = "#e15759", "0" = "#f5f5f5", "1" = "#59a14f"),
    labels = c("Short", "Flat", "Long"),
    name = "Signal"
  ) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Sector Signals at Month-End Close (Training)",
       subtitle = "Green = long | White = flat | Execute at open of following month",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 30, hjust = 1, size = 8),
        panel.grid = element_blank(), legend.position = "top")

ggplotly(p_signals, tooltip = c("x", "y", "fill"))
```

## Training Drawdown

```{r appendix-training-drawdown}
dd_data <- training_perf %>%
  arrange(date) %>%
  mutate(
    peak_strat = cummax(cumeq),
    peak_spy   = cummax(cumeq_spy),
    peak_kf    = cummax(replace_na(cumeq_kf, 1)),
    dd_strat   = (cumeq     - peak_strat) / peak_strat,
    dd_spy     = (cumeq_spy - peak_spy)   / peak_spy,
    dd_strat_kf = (replace_na(cumeq_kf, 1) - peak_kf) / peak_kf
  )

mdd_strat_pt <- dd_data %>% slice_min(dd_strat, n = 1)
mdd_spy_pt   <- dd_data %>% slice_min(dd_spy,   n = 1)
mdd_kf_pt    <- dd_data %>% slice_min(dd_strat_kf, n = 1)

plot_ly(dd_data, x = ~date) %>%
  add_lines(y = ~dd_strat_kf, name = "Strategy + KF scaling",
            line = list(color = "#59a14f", width = 1.5)) %>%
  add_lines(y = ~dd_strat, name = "Strategy (net TC)",
            line = list(color = "#4e79a7", width = 1.5)) %>%
  add_lines(y = ~dd_spy, name = "SPY Buy & Hold",
            line = list(color = "#e15759", width = 1.5)) %>%
  add_markers(data = mdd_kf_pt, x = ~date, y = ~dd_strat_kf,
              name = paste0("KF MDD: ", scales::percent(mdd_kf_pt$dd_strat_kf, accuracy = 0.1)),
              marker = list(color = "#59a14f", size = 10, symbol = "circle")) %>%
  add_markers(data = mdd_spy_pt, x = ~date, y = ~dd_spy,
              name = paste0("SPY MDD: ", scales::percent(mdd_spy_pt$dd_spy, accuracy = 0.1)),
              marker = list(color = "#e15759", size = 10, symbol = "circle")) %>%
  layout(
    title = list(text = "Drawdown from Peak -- Training Period", font = list(size = 14)),
    xaxis = list(title = NULL, type = "date"),
    yaxis = list(title = "Drawdown from Peak", tickformat = ".0%", zeroline = TRUE),
    hovermode = "x unified",
    legend = list(x = 0.99, y = 0.01, xanchor = "right", yanchor = "bottom",
                  bgcolor = "rgba(255,255,255,0.8)")
  )
```

## Training Performance Summary

```{r appendix-training-stats-gt}
# Compute benchmark stats for 5050, defensive, cyclical
bench_5050_rets <- replace_na(bench_5050$ret, 0)
bench_def_rets  <- replace_na(bench_def$ret, 0)
bench_cyc_rets  <- replace_na(bench_cyc$ret, 0)

bench_5050_total <- last(training_perf$cumeq_5050, na_rm = TRUE) - 1
bench_def_total  <- last(training_perf$cumeq_def,  na_rm = TRUE) - 1
bench_cyc_total  <- last(training_perf$cumeq_cyc,  na_rm = TRUE) - 1

tibble(
  Metric = c("Total Return", "CAGR", "Ann. Volatility", "Sharpe Ratio", "Omega (L=0)",
             "Max Drawdown", "MDD Date"),
  `Strategy + KF` = c(
    scales::percent(last(training_perf$cumeq_kf, na.rm = TRUE) - 1, accuracy = 0.1),
    scales::percent((last(training_perf$cumeq_kf, na.rm = TRUE))^(1/n_years) - 1, accuracy = 0.1),
    "--",
    round(sharpe_kf, 3),
    round(omega_kf, 3),
    scales::percent(max_dd_kf, accuracy = 0.1),
    format(mdd_date_kf, "%Y-%m-%d")
  ),
  `SPY Buy & Hold` = c(
    scales::percent(total_spy, accuracy = 0.1),
    scales::percent(cagr_spy, accuracy = 0.1),
    scales::percent(ann_vol_spy, accuracy = 0.1),
    round(sharpe_spy, 3),
    round(omega_spy, 3),
    scales::percent(max_dd_spy, accuracy = 0.1),
    format(mdd_date_spy, "%Y-%m-%d")
  ),
  `50/50 SPY+Top3` = c(
    scales::percent(bench_5050_total, accuracy = 0.1),
    scales::percent((1 + bench_5050_total)^(1/n_years) - 1, accuracy = 0.1),
    scales::percent(sd(bench_5050_rets, na.rm = TRUE) * sqrt(252), accuracy = 0.1),
    round((mean(bench_5050_rets, na.rm = TRUE) / sd(bench_5050_rets, na.rm = TRUE)) * sqrt(252), 3),
    round(omega_fn(bench_5050_rets), 3),
    scales::percent(min((training_perf$cumeq_5050 - cummax(training_perf$cumeq_5050)) / cummax(training_perf$cumeq_5050), na.rm = TRUE), accuracy = 0.1),
    "--"
  ),
  `100% Defensive` = c(
    scales::percent(bench_def_total, accuracy = 0.1),
    scales::percent((1 + bench_def_total)^(1/n_years) - 1, accuracy = 0.1),
    scales::percent(sd(bench_def_rets, na.rm = TRUE) * sqrt(252), accuracy = 0.1),
    round((mean(bench_def_rets, na.rm = TRUE) / sd(bench_def_rets, na.rm = TRUE)) * sqrt(252), 3),
    round(omega_fn(bench_def_rets), 3),
    scales::percent(min((training_perf$cumeq_def - cummax(training_perf$cumeq_def)) / cummax(training_perf$cumeq_def), na.rm = TRUE), accuracy = 0.1),
    "--"
  ),
  `100% Cyclical` = c(
    scales::percent(bench_cyc_total, accuracy = 0.1),
    scales::percent((1 + bench_cyc_total)^(1/n_years) - 1, accuracy = 0.1),
    scales::percent(sd(bench_cyc_rets, na.rm = TRUE) * sqrt(252), accuracy = 0.1),
    round((mean(bench_cyc_rets, na.rm = TRUE) / sd(bench_cyc_rets, na.rm = TRUE)) * sqrt(252), 3),
    round(omega_fn(bench_cyc_rets), 3),
    scales::percent(min((training_perf$cumeq_cyc - cummax(training_perf$cumeq_cyc)) / cummax(training_perf$cumeq_cyc), na.rm = TRUE), accuracy = 0.1),
    "--"
  )
) %>%
  gt() %>%
  tab_header(
    title = "Training Period Performance Summary",
    subtitle = "Jun 2000 -- Dec 2024 | 5 bps/side TC | Flat periods earn rf_daily (^IRX)"
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#f8f9fa"), cell_text(weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Metric)
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    heading.subtitle.font.size = px(13),
    table.border.top.color = "#4e79a7",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "#4e79a7"
  )
```

## Test Period Drawdown

```{r appendix-test-drawdown}
test_dd <- test_perf %>%
  arrange(date) %>%
  mutate(peak_strat = cummax(cumeq),
         peak_spy   = cummax(cumeq_spy),
         dd_strat   = (cumeq     - peak_strat) / peak_strat,
         dd_spy     = (cumeq_spy - peak_spy)   / peak_spy)

mdd_strat_test <- test_dd %>% slice_min(dd_strat, n = 1)
mdd_spy_test   <- test_dd %>% slice_min(dd_spy,   n = 1)

plot_ly(test_dd, x = ~date) %>%
  add_lines(y = ~dd_strat, name = "Strategy",
            line = list(color = "#59a14f", width = 1.5)) %>%
  add_lines(y = ~dd_spy, name = "SPY Buy & Hold",
            line = list(color = "#e15759", width = 1.5)) %>%
  add_markers(data = mdd_strat_test, x = ~date, y = ~dd_strat,
              name = paste0("Strategy MDD: ", scales::percent(mdd_strat_test$dd_strat, accuracy = 0.1)),
              marker = list(color = "#59a14f", size = 10, symbol = "circle")) %>%
  add_markers(data = mdd_spy_test, x = ~date, y = ~dd_spy,
              name = paste0("SPY MDD: ", scales::percent(mdd_spy_test$dd_spy, accuracy = 0.1)),
              marker = list(color = "#e15759", size = 10, symbol = "circle")) %>%
  layout(
    title = list(text = "Test Period Drawdown from Peak", font = list(size = 14)),
    xaxis = list(title = NULL, type = "date"),
    yaxis = list(title = "Drawdown", tickformat = ".0%"),
    hovermode = "x unified",
    legend = list(x = 0.99, y = 0.01, xanchor = "right", yanchor = "bottom",
                  bgcolor = "rgba(255,255,255,0.8)")
  )
```

## Test Period Regime Assignment

```{r appendix-test-regime}
#| fig-height: 3
test_regime_ts <- test_signals %>%
  distinct(rebal_date, regime) %>%
  mutate(regime = factor(regime))

ggplot(test_regime_ts, aes(x = rebal_date, y = 1, fill = regime)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = c("1" = "#4e79a7", "2" = "#f28e2b", "3" = "#59a14f"),
    na.value = "grey85", name = "Regime"
  ) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(title = "Layer 1: Regime Assignment -- Test Period",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid = element_blank(), legend.position = "top")
```

## Test Period Monthly Diagnostics

```{r appendix-test-monthly-diag}
test_monthly_diag <- test_signals %>%
  select(rebal_date, ticker, signal, avg_ret, regime, sector_cluster) %>%
  group_by(rebal_date) %>%
  summarise(
    n_long  = sum(signal == 1L),
    n_flat  = sum(signal == 0L),
    longs   = paste(ticker[signal == 1L], collapse = ", "),
    regime  = first(regime),
    .groups = "drop"
  ) %>%
  left_join(features_wide %>% select(date, vix), by = c("rebal_date" = "date")) %>%
  left_join(
    test_pnl2 %>%
      mutate(ym = floor_date(date, "month")) %>%
      group_by(ym) %>%
      summarise(month_ret = prod(1 + ret) - 1, .groups = "drop") %>%
      mutate(rebal_date = ym + months(1) - days(1)),
    by = "rebal_date"
  ) %>%
  select(rebal_date, regime, vix, n_long, longs, month_ret)

test_monthly_diag %>%
  mutate(
    rebal_date = format(rebal_date, "%Y-%m-%d"),
    vix = round(vix, 1)
  ) %>%
  gt() %>%
  tab_header(title = "Test Period Monthly Diagnostics") %>%
  fmt_percent(columns = month_ret, decimals = 1) %>%
  cols_label(
    rebal_date = "Rebal Date", regime = "Regime", vix = "VIX",
    n_long = "# Long", longs = "Long Sectors", month_ret = "Month Return"
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#f8f9fa"), cell_text(weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    table.border.top.color = "#4e79a7",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "#4e79a7"
  )
```

