---
title: "FIN 452 — Quantitative Trading Strategy"
subtitle: "Regime-Based Sector Rotation via Clustering & Kalman Filter"
author: "Bao Nguyen"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-width: 12
    fig-height: 8
execute:
  warning: false
  message: false
  echo: true
  cache: false
---

```{r setup}
library(tidyverse)
library(lubridate)
library(tidyquant)
library(tidymodels)
library(tidyclust)
library(timetk)
library(TTR)
library(PerformanceAnalytics)
library(xts)
library(zoo)
library(quantmod)
library(RTL)
library(lattice)
library(plotly)
library(foreach)
library(doParallel)
library(dlm)
library(cluster)
library(factoextra)
library(patchwork)
library(ggrepel)

# Helper: rolling function with partial window support
# Computes as soon as min_obs are available; doesn't wait for full window.
# Allows first-month ETF data to produce values without the full lookback.
roll_partial <- function(x, width, FUN, min_obs = 5, ...) {
  rollapply(x, width = width, FUN = function(z) {
    z_clean <- z[!is.na(z)]
    if (length(z_clean) < min_obs) return(NA_real_)
    FUN(z_clean, ...)
  }, fill = NA, align = "right", partial = TRUE)
}

# Helper: Kalman filter for a single indicator series (local level model)
# Fits MLE observation noise (V) and process noise (W), then applies causal
# dlmFilter (forward-pass only — no lookahead). NAs treated as missing
# observations; filter propagates prior without updating.
kf_level <- function(x, min_obs = 30) {
  out <- rep(NA_real_, length(x))
  if (sum(!is.na(x)) < min_obs) return(out)
  fit <- tryCatch(
    dlmMLE(x, parm = c(0, 0),
           build = function(p) dlmModPoly(1, dV = exp(p[1]), dW = exp(p[2]))),
    error = function(e) NULL
  )
  if (is.null(fit) || fit$convergence != 0) return(out)
  mod  <- dlmModPoly(1, dV = exp(fit$par[1]), dW = exp(fit$par[2]))
  filt <- dlmFilter(x, mod)
  drop(filt$m[-1])   # m is (n+1): first row is prior m_0, drop it
}

# Helper: rolling beta (two-series covariance / variance)
roll_beta <- function(sector_ret, spy_ret, width = 252, min_obs = 20) {
  mat <- cbind(sector_ret, spy_ret)
  rollapply(mat, width = width, FUN = function(x) {
    cc <- complete.cases(x)
    if (sum(cc) < min_obs) return(NA_real_)
    cov(x[cc, 1], x[cc, 2]) / var(x[cc, 2])
  }, by.column = FALSE, fill = NA, align = "right", partial = TRUE)
}
```

## Universe & Periods

```{r params}
tickers <- c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
             "XLK", "XLP", "XLRE", "XLU", "XLV", "XLY")

# Availability constraints
etf_inception <- tribble(
  ~ticker, ~inception,
  "SPY",   "1993-01-29",
  "XLB",   "1998-12-22",
  "XLC",   "2018-06-18",
  "XLE",   "1998-12-22",
  "XLF",   "1998-12-22",
  "XLI",   "1998-12-22",
  "XLK",   "1998-12-22",
  "XLP",   "1998-12-22",
  "XLRE",  "2015-10-07",
  "XLU",   "1998-12-22",
  "XLV",   "1998-12-22",
  "XLY",   "1998-12-22"
) %>% mutate(inception = as.Date(inception))

TRAIN_START <- as.Date("2000-06-01")
TRAIN_END   <- as.Date("2024-12-31")
TEST_START  <- as.Date("2025-01-01")
TEST_END    <- as.Date("2026-03-31")

PULL_FROM   <- as.Date("1998-01-01")   # pull full history; filter later
```

## Data Loading

```{r load-data}
#| cache: true

# ⚠️ COURSE PATTERN: tq_get → toTitleCase → xts → adjustOHLC → tibble
load_etf <- function(ticker) {
  raw <- tryCatch(
    tidyquant::tq_get(ticker, from = format(PULL_FROM)),
    error = function(e) NULL
  )
  if (is.null(raw) || isFALSE(raw) || nrow(raw) == 0) {
    warning(paste("tq_get failed for:", ticker))
    return(NULL)
  }
  raw %>%
    dplyr::rename_all(tools::toTitleCase) %>%
    timetk::tk_xts(date_var = Date) %>%
    quantmod::adjustOHLC(use.Adjusted = TRUE) %>%
    timetk::tk_tbl(rename_index = "Date") %>%
    dplyr::select(-Adjusted) %>%
    dplyr::mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
    dplyr::rename(date = Date) %>%
    dplyr::mutate(ticker = ticker)
}

raw_data <- tickers %>%
  purrr::map(load_etf) %>%
  purrr::compact() %>%          # drop any NULL returns from failed downloads
  dplyr::bind_rows() %>%
  left_join(etf_inception, by = "ticker") %>%
  filter(date >= inception) %>%        # drop pre-inception rows
  select(-inception) %>%
  arrange(ticker, date)

# Training and test subsets
train_data <- raw_data %>% filter(date >= TRAIN_START, date <= TRAIN_END)
test_data  <- raw_data %>% filter(date >= TEST_START,  date <= TEST_END)

glimpse(raw_data)
```

## Interest Rates

All rate data pulled via `tq_get()` — no API key required. 10Y yield (`^TNX`) and 3M T-bill (`^IRX`) from Yahoo Finance; 2Y yield (`DGS2`) and 10Y breakeven inflation (`T10YIE`) from FRED via `get = "economic.data"` (public HTTP, no authentication). Yield curve slope = 10Y − 2Y. Breakeven inflation captures the inflation regime dimension — orthogonal to the yield curve slope and VIX.

3M T-bill rate (`^IRX`) is pulled separately as the daily risk-free rate for Sharpe ratio calculation in `tradeStats()`. It is **not** used as a clustering feature.

```{r load-rates}
#| cache: true

rates_raw <- bind_rows(
  tq_get("^TNX", from = format(PULL_FROM)) %>%
    transmute(date, series = "y10", value = close),
  tq_get("DGS2", get = "economic.data", from = format(PULL_FROM)) %>%
    transmute(date, series = "y2", value = price),
  tq_get("T10YIE", get = "economic.data", from = format(PULL_FROM)) %>%
    transmute(date, series = "breakeven", value = price)
) %>%
  pivot_wider(names_from = series, values_from = value)

# Yield curve slope: 10Y minus 2Y (percentage points)
# Positive = normal curve; negative = inverted (recession signal)
# Breakeven inflation: market-implied 10Y avg inflation (orthogonal regime dimension)
rates <- rates_raw %>%
  mutate(yield_curve = y10 - y2) %>%
  filter(date >= TRAIN_START)

# Risk-free rate (3M T-bill) for Sharpe calculation — NOT a clustering feature
# Annualised % → daily decimal (divide by 100 for %, then by 252 for daily)
rf_daily <- tq_get("^IRX", from = format(TRAIN_START)) %>%
  transmute(date, rf_daily = close / 100 / 252)
```

```{r chart-rates}
rates_long <- rates %>%
  pivot_longer(-date, names_to = "series", values_to = "value") %>%
  mutate(series = recode(series,
    y10        = "10Y Treasury",
    y2         = "2Y Treasury",
    yield_curve = "Yield Curve (10Y − 2Y)",
    breakeven  = "Breakeven Inflation (T10YIE)"))

p_rates <- ggplot(rates_long, aes(x = date, y = value, colour = series)) +

  annotate("rect", xmin = TEST_START,
           xmax = max(rates_long$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +

  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.4) +

  geom_hline(yintercept = 0, linetype = "dotted",
             colour = "grey50", linewidth = 0.3) +

  geom_line(linewidth = 0.5, alpha = 0.85) +

  scale_colour_manual(values = c(
    "10Y Treasury"                  = "#4e79a7",
    "2Y Treasury"                   = "#76b7b2",
    "Yield Curve (10Y − 2Y)"        = "#e15759",
    "Breakeven Inflation (T10YIE)"  = "#f28e2b"
  )) +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +

  labs(
    title    = "Treasury Yields, Yield Curve & Breakeven Inflation",
    subtitle = "Inverted curve (red below zero) = recession signal | Orange = inflation regime",
    x = NULL, y = "Rate (%)", colour = NULL
  ) +

  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))

ggplotly(p_rates, tooltip = c("x", "y", "colour")) %>%
  layout(legend = list(orientation = "h", y = 1.08))
```

## Historical Price Levels

Each ETF indexed to 100 at its first available date within the chart window. Vertical line marks the train/test split. Greyed region = out-of-sample test period.

```{r chart-levels}
# Index each ETF to 100 at its first observation, then pivot wide (one column per ticker)
display_data <- raw_data %>%
  filter(date >= TRAIN_START) %>%
  group_by(ticker) %>%
  arrange(date) %>%
  mutate(level = 100 * Close / first(Close)) %>%
  ungroup() %>%
  select(date, ticker, level) %>%
  pivot_wider(names_from = ticker, values_from = level)

# Pivot back to long for plotting
p <- display_data %>%
  pivot_longer(-date, names_to = "ticker", values_to = "level") %>%
  mutate(ticker = factor(ticker, levels = tickers)) %>%
  ggplot(aes(x = date, y = level)) +

  # Test period shading
  annotate("rect",
           xmin = TEST_START, xmax = TEST_END,
           ymin = -Inf, ymax = Inf,
           fill = "#f0f0f0", alpha = 0.8) +

  # Train/test split line
  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.4) +

  geom_line(colour = "steelblue", linewidth = 0.4) +

  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +

  annotate("text",
           x = TEST_START + 30, y = -Inf,
           label = "Test", hjust = 0, vjust = -0.5,
           size = 2.5, colour = "grey40") +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(suffix = "")) +

  labs(
    title    = "SPDR Sector ETF Historical Levels (Indexed to 100 at Training Start)",
    subtitle = "Training: Jun 2000 – Dec 2024  |  Test: Jan 2025 – Mar 2026  |  * = later inception date",
    x        = NULL,
    y        = "Indexed Level (100 = first observation)"
  ) +

  theme_minimal(base_size = 10) +
  theme(
    strip.text       = element_text(size = 7.5, face = "bold"),
    axis.text.x      = element_text(size = 7, angle = 30, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )

ggplotly(p, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

## Rolling Volatility

### Compute

```{r compute-features}
#| cache: true

# Pull VIX — already annualised % (e.g. 20 = 20% annual vol), no transformation needed
vix_raw <- tidyquant::tq_get("^VIX", from = format(PULL_FROM)) %>%
  select(date, vix = close)

# Get SPY returns separately for beta computation
spy_returns <- raw_data %>%
  filter(ticker == "SPY", date >= TRAIN_START) %>%
  arrange(date) %>%
  mutate(spy_retClCl = Close / lag(Close) - 1,
         spy_log_ret = log(Close / lag(Close))) %>%
  select(date, spy_retClCl, spy_log_ret)

# 21 trading days = 30 calendar days * 252/365 ≈ 21
# This matches VIX's 30-calendar-day horizon for apples-to-apples VRP
RVOL_WINDOW  <- 21
BETA_WINDOW  <- 252
ZSCORE_WINDOW <- 252

# Compute all rolling features per ETF
# partial = TRUE via roll_partial() allows first-month data to produce values
features_long <- raw_data %>%
  filter(date >= TRAIN_START) %>%
  group_by(ticker) %>%
  arrange(date) %>%
  mutate(
    retClCl     = Close / lag(Close) - 1,
    log_ret     = log(Close / lag(Close)),

    # 21d rolling vol (annualised %)
    rvol_21     = roll_partial(retClCl, width = RVOL_WINDOW, FUN = sd,
                               min_obs = 5) * sqrt(252) * 100,

    # 21d rolling log return (sum of daily log returns)
    # Partial window: first-month ETFs get whatever days are available
    log_ret_21  = roll_partial(log_ret, width = RVOL_WINDOW, FUN = sum,
                               min_obs = 1)
  ) %>%
  ungroup() %>%
  # Join SPY returns for beta computation
  left_join(spy_returns, by = "date") %>%
  # Join VIX for spread and VRP
  left_join(vix_raw, by = "date") %>%
  group_by(ticker) %>%
  arrange(date) %>%
  mutate(
    # 252d rolling beta to SPY (partial: starts computing at 20 obs)
    beta_252 = roll_beta(retClCl, spy_retClCl, width = BETA_WINDOW, min_obs = 20),

    # Vol spread: sector realised vol minus VIX
    # Levels are biased by sector character — z-score removes this
    spread   = rvol_21 - vix,

    # 252d rolling z-score of the spread
    # Removes sector-specific mean AND variance → directly comparable across sectors
    spread_z = {
      sp_mean <- roll_partial(spread, width = ZSCORE_WINDOW, FUN = mean, min_obs = 30)
      sp_sd   <- roll_partial(spread, width = ZSCORE_WINDOW, FUN = sd,   min_obs = 30)
      (spread - sp_mean) / sp_sd
    }
  ) %>%
  ungroup() %>%
  select(date, ticker, retClCl, log_ret, rvol_21, log_ret_21,
         beta_252, vix, spread, spread_z)

# Relative features: sector vs SPY benchmark
# SPY row exists in features_long — pull it as reference series
spy_ref <- features_long %>%
  filter(ticker == "SPY") %>%
  select(date, spy_rvol_21 = rvol_21, spy_log_ret_21 = log_ret_21)

features_long <- features_long %>%
  left_join(spy_ref, by = "date") %>%
  mutate(
    # Excess vol: how much more/less volatile than SPY (not VIX)
    rel_vol      = rvol_21 - spy_rvol_21,
    # Excess momentum: sector return minus market return — isolates alpha
    rel_momentum = log_ret_21 - spy_log_ret_21
  ) %>%
  select(-spy_rvol_21, -spy_log_ret_21)

# Wide format: one row per date, metrics as columns per ticker
features_wide <- features_long %>%
  select(date, ticker, rvol_21, spread, spread_z, log_ret_21, beta_252,
         rel_vol, rel_momentum) %>%
  pivot_wider(names_from = ticker,
              values_from = c(rvol_21, spread, spread_z, log_ret_21, beta_252,
                              rel_vol, rel_momentum),
              names_glue = "{ticker}_{.value}") %>%
  left_join(vix_raw, by = "date") %>%
  left_join(rates %>% select(date, yield_curve, y10, y2, breakeven), by = "date") %>%
  mutate(vrp = vix - SPY_rvol_21)   # VRP: apples-to-apples, both 21 trading days
```

## Kalman Filter

Each clustering input is smoothed independently with a local level model (random walk + noise). MLE estimates observation noise V and process noise W per series. Uses `dlmFilter()` only — causal forward pass, no lookahead.

Columns smoothed: `vix`, `vrp`, `yield_curve`, `breakeven`, and all per-sector `_spread_z`, `_beta_252`, `_log_ret_21`, `_rel_vol`, `_rel_momentum` columns. Raw OHLCV-derived columns (`rvol_21`, `spread`, rate levels) are retained but not filtered — they are not clustering inputs.

```{r kalman-filter}
# Identify columns to filter (clustering inputs only)
kf_cols <- c(
  "vix", "vrp", "yield_curve", "breakeven",
  grep("_spread_z$|_beta_252$|_log_ret_21$|_rel_vol$|_rel_momentum$",
       names(features_wide), value = TRUE)
)

# Apply kf_level independently to each clustering column
# All other columns (dates, raw vols, rate levels) pass through unchanged
features_kf <- features_wide %>%
  arrange(date) %>%
  mutate(across(all_of(kf_cols), kf_level))
```

### SPY Rolling Vol vs VIX

21-trading-day realised vol for SPY plotted against VIX. Both cover ~30 calendar days (21 trading days = 30 * 252/365) — directly comparable. The gap is the Volatility Risk Premium (VRP): positive = fear premium; negative = crash worse than anticipated.

```{r chart-spy-vol}
spy_vol <- features_wide %>%
  select(date, `21d Realised` = SPY_rvol_21, VIX = vix, VRP = vrp) %>%
  pivot_longer(-date, names_to = "series", values_to = "vol") %>%
  mutate(series = factor(series, levels = c("VIX", "21d Realised", "VRP")))

p_spy <- ggplot(spy_vol, aes(x = date, y = vol, colour = series)) +

  annotate("rect",
           xmin = TEST_START, xmax = max(spy_vol$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +

  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.4) +

  geom_hline(yintercept = 0, linetype = "dotted",
             colour = "grey60", linewidth = 0.3) +

  geom_line(linewidth = 0.5, alpha = 0.85) +

  scale_colour_manual(values = c(
    "VIX"          = "#e15759",
    "21d Realised" = "#4e79a7",
    "VRP"          = "#f28e2b"
  )) +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +

  labs(
    title    = "SPY: VIX vs 21d Realised Volatility & VRP",
    subtitle = "VRP = VIX − 21d Realised Vol (21 trading days ≈ 30 calendar days)  |  Positive = fear premium",
    x        = NULL, y = "Annualised Volatility (%)", colour = NULL
  ) +

  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "top",
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )

ggplotly(p_spy, tooltip = c("x", "y", "colour")) %>%
  layout(legend = list(orientation = "h", y = 1.08))
```

### Sector ETF Vol Spread Z-Score

Rolling z-score of each sector's vol spread (sector rvol − VIX), normalised by that sector's own 252-day rolling mean and standard deviation. Removes structural level differences — a z-score of +2 means the same anomalous stress regardless of whether the sector is normally above or below VIX. Partial window support: values appear earlier than 252 days but require at least 30 observations for the z-score.

```{r chart-sector-vol}
sector_tickers <- setdiff(tickers, "SPY")

sector_vol <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  mutate(ticker = factor(ticker, levels = sector_tickers))

p_sec <- ggplot(sector_vol, aes(x = date, y = spread_z)) +

  annotate("rect",
           xmin = TEST_START, xmax = max(sector_vol$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +

  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.3) +

  # Zero line = sector's historical norm
  geom_hline(yintercept = 0, linetype = "dotted",
             colour = "grey50", linewidth = 0.3) +

  geom_line(colour = "#4e79a7", linewidth = 0.35, alpha = 0.85) +

  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +

  labs(
    title    = "Sector Vol Spread Z-Score  (sector rvol − VIX, normalised)",
    subtitle = "Zero = historical norm  |  +2 = 2sd more stressed vs VIX than usual  |  Comparable across sectors",
    x        = NULL, y = "Z-Score (standard deviations)"
  ) +

  theme_minimal(base_size = 10) +
  theme(
    strip.text       = element_text(size = 8, face = "bold"),
    axis.text.x      = element_text(size = 7, angle = 30, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )

ggplotly(p_sec, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

## Rolling Beta to SPY

252-day rolling beta per sector vs SPY. Measures each sector's market sensitivity over time — a regime signal. When beta changes, the sector is coupling to or decoupling from the broad market. Partial window: starts producing values at 20 observations.

```{r chart-beta}
beta_data <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  mutate(ticker = factor(ticker, levels = sector_tickers))

p_beta <- ggplot(beta_data, aes(x = date, y = beta_252)) +

  annotate("rect",
           xmin = TEST_START, xmax = max(beta_data$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +

  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.3) +

  geom_hline(yintercept = 1, linetype = "dotted",
             colour = "grey50", linewidth = 0.3) +

  geom_line(colour = "#59a14f", linewidth = 0.35, alpha = 0.85) +

  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +

  labs(
    title    = "252d Rolling Beta to SPY",
    subtitle = "Beta = 1 line shown  |  Above 1 = amplifies SPY moves  |  Below 1 = dampens / decouples",
    x        = NULL, y = "Beta"
  ) +

  theme_minimal(base_size = 10) +
  theme(
    strip.text       = element_text(size = 8, face = "bold"),
    axis.text.x      = element_text(size = 7, angle = 30, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )

ggplotly(p_beta, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

## 21d Rolling Log Return

Cumulative log return over 21 trading days (~1 month). Captures short-term sector momentum. Partial window: first-month ETFs produce values from whatever days are available.

```{r chart-logret}
logret_data <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  mutate(ticker = factor(ticker, levels = sector_tickers))

p_logret <- ggplot(logret_data, aes(x = date, y = log_ret_21)) +

  annotate("rect",
           xmin = TEST_START, xmax = max(logret_data$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +

  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.3) +

  geom_hline(yintercept = 0, linetype = "dotted",
             colour = "grey50", linewidth = 0.3) +

  geom_line(colour = "#b07aa1", linewidth = 0.35, alpha = 0.85) +

  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_percent()) +

  labs(
    title    = "21d Rolling Log Return (Sector Momentum)",
    subtitle = "Cumulative log return over 21 trading days  |  Positive = upward momentum  |  Negative = downward",
    x        = NULL, y = "Log Return"
  ) +

  theme_minimal(base_size = 10) +
  theme(
    strip.text       = element_text(size = 8, face = "bold"),
    axis.text.x      = element_text(size = 7, angle = 30, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )

ggplotly(p_logret, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

---

# Clustering

## Feature Matrix Preparation

Clustering inputs are the Kalman-filtered macro features (`vix`, `vrp`, `yield_curve`, `breakeven`) plus per-sector `_spread_z`, `_beta_252`, and `_log_ret_21` for the 11 sector ETFs (SPY excluded — its beta to itself = 1, its spread_z ≈ −VRP). Rows with any NA are dropped; SPY-derived macro features already cover SPY's contribution.

```{r cluster-prep}
# Clustering features: macro + per-sector (sectors only, not SPY)
cluster_cols <- c(
  "vix", "vrp", "yield_curve", "breakeven",
  paste0(sector_tickers, "_spread_z"),
  paste0(sector_tickers, "_beta_252"),
  paste0(sector_tickers, "_log_ret_21"),
  paste0(sector_tickers, "_rel_vol"),
  paste0(sector_tickers, "_rel_momentum")
)

# Training-period rows only; drop any date with incomplete features
cluster_input <- features_kf %>%
  filter(date >= TRAIN_START, date <= TRAIN_END) %>%
  select(date, all_of(cluster_cols)) %>%
  drop_na()

# Scaled matrix for distance-based methods (hierarchical needs this too)
cluster_recipe <- recipe(~ ., data = cluster_input %>% select(-date)) %>%
  step_normalize(all_predictors())

cluster_prep   <- prep(cluster_recipe)
cluster_mat    <- bake(cluster_prep, new_data = NULL) %>% as.matrix()
```

## K Selection Diagnostics

Elbow plot (total within-cluster sum of squares) and average silhouette width for K = 2–6. The elbow shows where additional clusters stop reducing compactness meaningfully; silhouette peaks at the K with best-separated clusters. Set `K_CLUSTERS` in the next chunk accordingly.

```{r cluster-k-select}
set.seed(42)

# --- K-means: WSS and silhouette for K = 2:6 ---
k_metrics <- map_dfr(2:6, function(k) {
  km  <- kmeans(cluster_mat, centers = k, nstart = 25, iter.max = 200)
  sil <- cluster::silhouette(km$cluster, dist(cluster_mat))
  tibble(k = k, wss = km$tot.withinss, silhouette = mean(sil[, 3]))
})

p_elbow <- ggplot(k_metrics, aes(x = k, y = wss)) +
  geom_line(colour = "#4e79a7", linewidth = 0.8) +
  geom_point(colour = "#4e79a7", size = 3) +
  scale_x_continuous(breaks = 2:6) +
  labs(title = "K-Means Elbow Plot", x = "Number of Clusters (K)",
       y = "Total Within-Cluster SS") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

p_sil <- ggplot(k_metrics, aes(x = k, y = silhouette)) +
  geom_line(colour = "#e15759", linewidth = 0.8) +
  geom_point(colour = "#e15759", size = 3) +
  geom_vline(xintercept = k_metrics$k[which.max(k_metrics$silhouette)],
             linetype = "dashed", colour = "grey50") +
  scale_x_continuous(breaks = 2:6) +
  labs(title = "Average Silhouette Width", x = "Number of Clusters (K)",
       y = "Mean Silhouette") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

p_elbow + p_sil
```

## Cluster Fit

`K_CLUSTERS` is set from the K selection diagnostics above — change it here to propagate everywhere.

```{r cluster-fit}
K_CLUSTERS <- 4   # <-- adjust after inspecting elbow/silhouette

set.seed(42)

km_spec <- k_means(num_clusters = K_CLUSTERS) %>% set_engine("stats")

km_wf  <- workflow() %>%
  add_recipe(cluster_recipe) %>%
  add_model(km_spec)

km_fit <- fit(km_wf, data = cluster_input %>% select(-date))

cluster_assignments <- augment(km_fit, new_data = cluster_input) %>%
  rename(km_cluster = .pred_cluster) %>%
  mutate(km_cluster = as.integer(str_remove(km_cluster, "Cluster_")))
```

## PCA Biplot

PCA reduces the 37-feature space to 2 dimensions. Points are coloured by cluster assignment. Well-separated blobs = stable, distinct regimes.

```{r cluster-pca}
pca_res  <- prcomp(cluster_mat, center = FALSE, scale. = FALSE)  # already normalised
pca_vals <- as_tibble(pca_res$x[, 1:2]) %>%
  bind_cols(cluster_assignments %>% select(date, km_cluster)) %>%
  mutate(km_cluster = as.factor(km_cluster))

pct_var <- round(summary(pca_res)$importance[2, 1:2] * 100, 1)

p_pca <- ggplot(pca_vals, aes(x = PC1, y = PC2, colour = km_cluster)) +
  geom_point(alpha = 0.4, size = 1.0) +
  scale_colour_manual(values = c("#4e79a7","#e15759","#59a14f","#f28e2b",
                                  "#b07aa1","#76b7b2")[seq_len(K_CLUSTERS)]) +
  labs(
    title    = "K-Means Clusters — PCA Biplot",
    subtitle = paste0("PC1 ", pct_var[1], "% | PC2 ", pct_var[2], "% of variance"),
    x = paste0("PC1 (", pct_var[1], "%)"),
    y = paste0("PC2 (", pct_var[2], "%)"),
    colour = "Cluster"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

ggplotly(p_pca, tooltip = c("x", "y", "colour"))
```

## Cluster Profiles

Mean feature values per K-means cluster. Standardised (z-score) so all features are on the same scale — positive = above average, negative = below. This is the economic regime interpretation step.

```{r cluster-profiles}
profile_data <- cluster_mat %>%
  as_tibble() %>%
  mutate(km_cluster = factor(cluster_assignments$km_cluster)) %>%
  pivot_longer(-km_cluster, names_to = "feature", values_to = "value") %>%
  group_by(km_cluster, feature) %>%
  summarise(mean_z = mean(value, na.rm = TRUE), .groups = "drop")

# Heatmap: features × clusters
p_profile <- ggplot(profile_data, aes(x = km_cluster, y = feature, fill = mean_z)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = round(mean_z, 2)), size = 2.5) +
  scale_fill_gradient2(low = "#4e79a7", mid = "white", high = "#e15759",
                       midpoint = 0, name = "Mean Z") +
  labs(
    title    = "Cluster Profiles — Mean Standardised Feature Values",
    subtitle = "Red = above average  |  Blue = below average  |  Interpret economically before labelling",
    x = "K-Means Cluster", y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(plot.title  = element_text(face = "bold"),
        axis.text.y = element_text(size = 7))

p_profile
```

---

# Signal Generation

## Parameters

```{r signal-params}
N_LONG           <- 3   # sectors to go long each month
N_SHORT          <- 0   # sectors to go short (0 = long-only)
MIN_CLUSTER_OBS  <- 6   # minimum cluster members before signal fires; else flat
VIX_THRESHOLD    <- 25  # go fully flat when raw VIX exceeds this level
DD_STOP          <- -0.20  # go flat next month if month-end drawdown exceeds this
```

## Expanding Window Signal Loop

At each month-end rebalancing date, K-means is refit on all data up to that date. The current date's cluster is identified, sector average returns within that cluster are ranked, and signals are generated. Cluster labels are not carried across refits — only the economic content (centroid values) matters.

```{r signal-generation}

# Sector daily returns for ranking within clusters
sector_ret <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  select(date, ticker, retClCl)

# Month-end rebalancing dates within training window
rebal_dates <- cluster_input %>%
  mutate(ym = lubridate::floor_date(date, "month")) %>%
  group_by(ym) %>%
  slice_max(date, n = 1) %>%
  ungroup() %>%
  arrange(date) %>%
  pull(date)

# Expanding window loop — refit K-means at each rebalancing date
signals_raw <- map_dfr(seq_along(rebal_dates), function(i) {
  t   <- rebal_dates[i]
  idx <- which(cluster_input$date == t)
  if (length(idx) == 0 || idx < K_CLUSTERS * 3) return(NULL)

  window_mat <- cluster_mat[1:idx, , drop = FALSE]

  set.seed(42 + i)   # reproducible but distinct per refit
  km <- kmeans(window_mat, centers = K_CLUSTERS, nstart = 25, iter.max = 200)

  current_cluster <- km$cluster[idx]
  cluster_dates   <- cluster_input$date[1:idx][km$cluster == current_cluster]

  # Helper: return a fully flat tibble for this date
  flat_signal <- function() {
    tibble(rebal_date = t, cluster = current_cluster,
           n_obs = length(cluster_dates),
           ticker = sector_tickers, avg_ret = NA_real_, signal = 0L)
  }

  # Flat if cluster history too thin
  if (length(cluster_dates) < MIN_CLUSTER_OBS) return(flat_signal())

  # VIX regime filter — go fully flat when market fear is elevated
  # Uses raw (non-KF) VIX so the filter reacts quickly to spikes
  current_vix <- features_wide %>% filter(date == t) %>% pull(vix)
  if (length(current_vix) == 1 && !is.na(current_vix) && current_vix > VIX_THRESHOLD) {
    return(flat_signal())
  }

  # Rank sectors by avg retClCl within current cluster
  rankings <- sector_ret %>%
    filter(date %in% cluster_dates) %>%
    group_by(ticker) %>%
    summarise(avg_ret = mean(retClCl, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(avg_ret)) %>%
    mutate(
      rank   = row_number(),
      n      = n(),
      signal = case_when(
        rank <= N_LONG & avg_ret > 0 ~ 1L,   # only long if cluster expects positive return
        rank > n - N_SHORT           ~ -1L,
        TRUE                         ~ 0L
      )
    )

  rankings %>%
    select(ticker, avg_ret, signal) %>%
    mutate(rebal_date = t, cluster = current_cluster, n_obs = length(cluster_dates))
})
```

## Cluster Assignment Over Time

Which regime each month-end was assigned to. Colour = cluster label within that refit (labels shift arbitrarily across refits — interpret by centroid, not by number).

```{r cluster-timeline}
cluster_timeline <- signals_raw %>%
  distinct(rebal_date, cluster, n_obs) %>%
  mutate(cluster = factor(cluster))

p_timeline <- ggplot(cluster_timeline,
                     aes(x = rebal_date, y = cluster, colour = cluster)) +
  geom_point(aes(size = n_obs), alpha = 0.7) +
  geom_line(aes(group = 1), alpha = 0.2, colour = "grey50") +
  scale_colour_manual(values = c("#4e79a7","#e15759","#59a14f","#f28e2b",
                                  "#b07aa1","#76b7b2")[seq_len(K_CLUSTERS)]) +
  scale_size_continuous(range = c(1, 5), name = "Cluster size") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title    = "Cluster Assignment at Each Month-End Rebalancing Date",
    subtitle = "Point size = number of historical dates in that cluster  |  Labels shift across refits — read centroid heatmap",
    x = NULL, y = "Cluster", colour = "Cluster"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")

ggplotly(p_timeline, tooltip = c("x", "y", "size"))
```

## Sector Signal Heatmap

Signals generated at each month-end close for execution at next month open. Green = long (+1), red = short (−1), white = flat (0). This is what the strategy holds during each month.

```{r signal-heatmap}
signal_heatmap <- signals_raw %>%
  mutate(ticker  = factor(ticker, levels = rev(sector_tickers)),
         signal  = factor(signal, levels = c(-1L, 0L, 1L)))

p_signals <- ggplot(signal_heatmap,
                    aes(x = rebal_date, y = ticker, fill = signal)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = c("-1" = "#e15759", "0" = "#f5f5f5", "1" = "#59a14f"),
    labels = c("Short (−1)", "Flat (0)", "Long (+1)")
  ) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title    = "Sector Signals Generated at Month-End Close",
    subtitle = "Execute at open of following month  |  Green = long  |  Red = short  |  White = flat",
    x = NULL, y = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold"),
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 8),
    panel.grid       = element_blank(),
    legend.position  = "top"
  )

ggplotly(p_signals, tooltip = c("x", "y", "fill"))
```

## Most Recent Signal

What the strategy signals for the most recent month in the training window — the last "read" before the test period.

```{r signal-latest}
signals_raw %>%
  filter(rebal_date == max(rebal_date)) %>%
  arrange(desc(avg_ret)) %>%
  mutate(
    signal_label = case_when(signal == 1L ~ "LONG", signal == -1L ~ "SHORT",
                             TRUE ~ "flat"),
    avg_ret_pct  = scales::percent(avg_ret, accuracy = 0.01)
  ) %>%
  select(rebal_date, ticker, avg_ret_pct, signal_label, n_obs) %>%
  knitr::kable(
    col.names = c("Rebal Date", "Sector", "Avg Cluster Return", "Signal", "Cluster Obs"),
    caption   = "Most recent month-end signal — execute at next month open"
  )
```

## End-of-Training Sector Snapshot

Market dynamics as of the last training date. Each sector is positioned by its Kalman-filtered relative momentum (x) and vol stress z-score (y). Point size = 252d beta to SPY. Colour = signal generated for the following month.

Quadrant interpretation:
- **Top-right**: high momentum + high stress — strong move but stretched
- **Bottom-right**: high momentum + low stress — clean uptrend, favourable
- **Top-left**: negative momentum + high stress — distressed, avoid or short
- **Bottom-left**: negative momentum + low stress — quiet underperformer

```{r sector-snapshot}
# Last available training date in features_kf
last_train_date <- features_kf %>%
  filter(date <= TRAIN_END) %>%
  pull(date) %>% max()

# Extract per-sector features at that date (pivot wide → long by sector)
sector_snapshot <- features_kf %>%
  filter(date == last_train_date) %>%
  select(date,
         matches("^(XL|SPY)[A-Z]*_rel_momentum$"),
         matches("^(XL|SPY)[A-Z]*_spread_z$"),
         matches("^(XL|SPY)[A-Z]*_beta_252$"),
         matches("^(XL|SPY)[A-Z]*_rel_vol$")) %>%
  pivot_longer(-date,
               names_to  = c("ticker", "metric"),
               names_pattern = "^(.+)_(rel_momentum|spread_z|beta_252|rel_vol)$") %>%
  filter(ticker %in% sector_tickers) %>%
  pivot_wider(names_from = metric, values_from = value)

# Join last signal
last_signal <- signals_raw %>%
  filter(rebal_date == max(rebal_date)) %>%
  select(ticker, signal, avg_ret)

sector_snapshot <- sector_snapshot %>%
  left_join(last_signal, by = "ticker") %>%
  mutate(
    signal_label = case_when(
      signal == 1L  ~ "Long",
      signal == -1L ~ "Short",
      TRUE          ~ "Flat"
    ),
    signal_label = factor(signal_label, levels = c("Long", "Flat", "Short"))
  )

p_snapshot <- ggplot(sector_snapshot,
                     aes(x = rel_momentum, y = spread_z,
                         colour = signal_label, size = beta_252,
                         label = ticker)) +
  geom_hline(yintercept =  1, linetype = "dotted", colour = "grey70") +
  geom_hline(yintercept = -1, linetype = "dotted", colour = "grey70") +
  geom_hline(yintercept =  0, linetype = "dashed",  colour = "grey50", linewidth = 0.4) +
  geom_vline(xintercept =  0, linetype = "dashed",  colour = "grey50", linewidth = 0.4) +
  geom_point(alpha = 0.85) +
  ggrepel::geom_text_repel(size = 3.5, show.legend = FALSE,
                            box.padding = 0.4, max.overlaps = 20) +
  scale_colour_manual(values = c("Long" = "#59a14f", "Flat" = "#bab0ac",
                                  "Short" = "#e15759")) +
  scale_size_continuous(range = c(3, 10), name = "Beta (252d)") +
  labs(
    title    = paste0("Sector Market Dynamics — End of Training (",
                      format(last_train_date, "%B %Y"), ")"),
    subtitle = "X = relative momentum vs SPY  |  Y = vol stress z-score vs norm  |  Size = beta to SPY",
    x        = "Relative Momentum (sector − SPY, 21d log return)",
    y        = "Vol Stress Z-Score (sector rvol − VIX, 252d normalised)",
    colour   = "Signal"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "right")

ggplotly(p_snapshot, tooltip = c("label", "x", "y", "size", "colour"))
```

## Animated Snapshot — All Month-Ends

Same chart as above but with a slider across every month-end rebalancing date in the training period. Use the play button or drag the slider to see how sector positions and signals evolved.

```{r sector-snapshot-animated}
# Raw (non-KF) features in long format — already has ticker column, no pivot needed
# Raw values show month-to-month variation; KF values are too smooth to animate usefully
snap_features <- features_long %>%
  filter(ticker %in% sector_tickers,
         date   %in% unique(signals_raw$rebal_date)) %>%
  select(date, ticker, spread_z, beta_252, rel_momentum, rel_vol)

# Join signals directly onto features — start from signals_raw to guarantee key alignment
all_snapshots <- signals_raw %>%
  select(rebal_date, ticker, signal) %>%
  left_join(snap_features, by = c("rebal_date" = "date", "ticker")) %>%
  mutate(
    signal_label = case_when(
      signal == 1L  ~ "Long",
      signal == -1L ~ "Short",
      TRUE          ~ "Flat"
    ),
    signal_label = factor(signal_label, levels = c("Long", "Flat", "Short")),
    frame_label  = format(rebal_date, "%Y-%m")
  ) %>%
  drop_na(rel_momentum, spread_z, beta_252) %>%
  mutate(bubble_size = scales::rescale(beta_252,
                                       to   = c(10, 30),
                                       from = c(0, max(abs(beta_252), na.rm = TRUE))))

# Compute fixed axis limits so axes don't jump between frames
x_range <- range(all_snapshots$rel_momentum, na.rm = TRUE) * 1.1
y_range <- range(all_snapshots$spread_z,     na.rm = TRUE) * 1.1

plot_ly(
  all_snapshots,
  x            = ~rel_momentum,
  y            = ~spread_z,
  size         = ~bubble_size,
  color        = ~signal_label,
  colors       = c("Long" = "#59a14f", "Flat" = "#bab0ac", "Short" = "#e15759"),
  text         = ~ticker,
  frame        = ~frame_label,
  type         = "scatter",
  mode         = "markers+text",
  textposition = "top center",
  marker       = list(sizemode = "diameter", sizeref = 1, opacity = 0.85)
) %>%
  animation_opts(frame = 600, transition = 300, redraw = FALSE) %>%
  animation_slider(currentvalue = list(prefix = "Month-end: ", font = list(size = 12))) %>%
  animation_button(x = 0, xanchor = "left", y = -0.12) %>%
  layout(
    title = list(text = "Sector Dynamics at Each Month-End Rebalancing Date",
                 font = list(size = 14)),
    xaxis = list(title = "Relative Momentum (sector − SPY, 21d)",
                 range = x_range, zeroline = TRUE, zerolinecolor = "#aaa"),
    yaxis = list(title = "Vol Stress Z-Score",
                 range = y_range, zeroline = TRUE, zerolinecolor = "#aaa"),
    legend = list(title = list(text = "Signal")),
    showlegend = TRUE
  )
```

## Animated Snapshot — Vol Stress vs Absolute Momentum

Same animation using `spread_z` (y) and raw `log_ret_21` (x) — absolute sector return rather than excess over SPY. During bull markets all sectors shift right together; during bear markets all shift left. This shows whether a sector's momentum is positive in absolute terms, while spread_z shows whether its vol is elevated relative to its own history.

```{r sector-snapshot-animated-2}
snap2 <- signals_raw %>%
  select(rebal_date, ticker, signal) %>%
  left_join(
    features_long %>%
      filter(ticker %in% sector_tickers,
             date   %in% unique(signals_raw$rebal_date)) %>%
      select(date, ticker, log_ret_21, spread_z, beta_252),
    by = c("rebal_date" = "date", "ticker")
  ) %>%
  mutate(
    signal_label = case_when(
      signal == 1L  ~ "Long",
      signal == -1L ~ "Short",
      TRUE          ~ "Flat"
    ),
    signal_label = factor(signal_label, levels = c("Long", "Flat", "Short")),
    frame_label  = format(rebal_date, "%Y-%m")
  ) %>%
  drop_na(log_ret_21, spread_z, beta_252) %>%
  mutate(bubble_size = scales::rescale(beta_252,
                                       to   = c(10, 30),
                                       from = c(0, max(abs(beta_252), na.rm = TRUE))))

x2_range <- range(snap2$log_ret_21, na.rm = TRUE) * 1.1
y2_range <- range(snap2$spread_z,   na.rm = TRUE) * 1.1

plot_ly(
  snap2,
  x            = ~log_ret_21,
  y            = ~spread_z,
  size         = ~bubble_size,
  color        = ~signal_label,
  colors       = c("Long" = "#59a14f", "Flat" = "#bab0ac", "Short" = "#e15759"),
  text         = ~ticker,
  frame        = ~frame_label,
  type         = "scatter",
  mode         = "markers+text",
  textposition = "top center",
  marker       = list(sizemode = "diameter", sizeref = 1, opacity = 0.85)
) %>%
  animation_opts(frame = 600, transition = 300, redraw = FALSE) %>%
  animation_slider(currentvalue = list(prefix = "Month-end: ", font = list(size = 12))) %>%
  animation_button(x = 0, xanchor = "left", y = -0.12) %>%
  layout(
    title = list(text = "Vol Stress Z-Score vs 21d Log Return",
                 font = list(size = 14)),
    xaxis = list(title = "21d Log Return (absolute)",
                 range = x2_range, zeroline = TRUE, zerolinecolor = "#aaa"),
    yaxis = list(title = "Vol Stress Z-Score",
                 range = y2_range, zeroline = TRUE, zerolinecolor = "#aaa"),
    legend = list(title = list(text = "Signal")),
    showlegend = TRUE
  )
```

---

# Training Period Performance

## P&L Calculation

Daily P&L chain for the training window. Signals are generated at each month-end close and executed at the open of the following trading day. Positions are held until the next rebalancing. Portfolio weight = 1/N_LONG per long sector; N_SHORT = 0 (long-only). Total allocation is always 100% when at least one long is active.

Transaction costs of **5 bps per side** (buy or sell) are applied only to positions that actually change at each rebalancing — sectors that are held across months pay no cost. Close-to-close daily returns are used throughout.

```{r training-pnl}
# -- Step 1: Map each rebal_date to its execution date (next available trading day) --
spy_dates <- raw_data %>%
  filter(ticker == "SPY", date >= TRAIN_START, date <= TRAIN_END) %>%
  arrange(date) %>%
  pull(date)

exec_map <- tibble(rebal_date = rebal_dates) %>%
  rowwise() %>%
  mutate(exec_date = {
    candidates <- spy_dates[spy_dates > rebal_date]
    if (length(candidates) == 0) NA_Date_ else candidates[1]
  }) %>%
  ungroup() %>%
  filter(!is.na(exec_date))

# -- Step 2: Build daily position series per sector --
# Start from exec_date each month, forward-fill until next exec_date
daily_signals <- signals_raw %>%
  select(rebal_date, ticker, signal) %>%
  inner_join(exec_map, by = "rebal_date") %>%
  select(date = exec_date, ticker, signal)

all_combos <- expand_grid(date = spy_dates, ticker = sector_tickers)

daily_pos <- all_combos %>%
  left_join(daily_signals, by = c("date", "ticker")) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  fill(signal, .direction = "down") %>%
  mutate(signal = replace_na(signal, 0L)) %>%   # flat before first signal fires
  ungroup()

# -- Step 3: Daily close-to-close returns per sector --
daily_rets <- raw_data %>%
  filter(ticker %in% sector_tickers, date %in% spy_dates) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(ret_cc = Close / dplyr::lag(Close) - 1) %>%
  ungroup() %>%
  select(date, ticker, ret_cc)

TC_BPS <- 0.0005   # 5 bps per side (buy or sell)

# Helper: compute daily portfolio weights from a position tibble
compute_weights <- function(pos_tbl) {
  pos_tbl %>%
    group_by(date) %>%
    mutate(
      n_long  = sum(signal == 1L,  na.rm = TRUE),
      n_short = sum(signal == -1L, na.rm = TRUE),
      weight  = case_when(
        signal ==  1L & n_long  > 0 ~  1 / n_long,
        signal == -1L & n_short > 0 ~ -1 / n_short,
        TRUE                        ~  0
      )
    ) %>%
    ungroup()
}

# -- Step 4a: Transaction costs on portfolio weight turnover --
# TC is proportional to how much weight actually moves, not signal integers.
# e.g. entering 1 of 3 longs = weight change of 1/3 → TC = 1/3 × 5bps, not 1 × 5bps.
# Self-financing is preserved: TC is funded from portfolio equity before compounding,
# so the investment base for the next period is reduced by the cost paid.
rebal_costs <- compute_weights(daily_pos) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(prev_weight = dplyr::lag(weight, default = 0)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(
    tc = sum(abs(weight - prev_weight), na.rm = TRUE) * TC_BPS,
    .groups = "drop"
  )

# -- Step 4b: First-pass P&L (VIX filter + avg_ret > 0 filter applied via signals_raw) --
daily_pnl <- compute_weights(daily_pos) %>%
  left_join(daily_rets, by = c("date", "ticker")) %>%
  group_by(date) %>%
  summarise(
    portfolio_ret_gross = sum(weight * replace_na(ret_cc, 0), na.rm = TRUE),
    n_long              = first(n_long),
    .groups             = "drop"
  ) %>%
  left_join(rebal_costs, by = "date") %>%
  mutate(
    tc            = replace_na(tc, 0),
    portfolio_ret = portfolio_ret_gross - tc,
    cumeq         = cumprod(1 + portfolio_ret),
    cumeq_gross   = cumprod(1 + portfolio_ret_gross)
  )

# -- Step 4c: Portfolio drawdown stop (second pass) --
# At each month-end, check the strategy's drawdown from its running peak.
# If drawdown < DD_STOP, go fully flat for the entire following month.
# This is causal: we observe drawdown at month-end before deciding next month's trades.
month_end_dd <- daily_pnl %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym) %>%
  slice_max(date, n = 1) %>%
  ungroup() %>%
  arrange(date) %>%
  mutate(
    peak    = cummax(cumeq),
    dd      = (cumeq - peak) / peak,
    in_stop = dd < DD_STOP,
    # Lead by 1 month: drawdown observed this month → next month is flat
    next_ym_stop = dplyr::lead(in_stop, default = FALSE)
  ) %>%
  mutate(ym = floor_date(date %m+% months(1), "month")) %>%   # next month key
  select(ym, next_ym_stop)

daily_pos_dd <- daily_pos %>%
  mutate(ym = floor_date(date, "month")) %>%
  left_join(month_end_dd, by = "ym") %>%
  mutate(
    next_ym_stop = replace_na(next_ym_stop, FALSE),
    signal       = if_else(next_ym_stop, 0L, signal)
  ) %>%
  select(-ym, -next_ym_stop)

# Second-pass TC and P&L with DD stop applied
rebal_costs_dd <- compute_weights(daily_pos_dd) %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(prev_weight = dplyr::lag(weight, default = 0)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(tc = sum(abs(weight - prev_weight), na.rm = TRUE) * TC_BPS, .groups = "drop")

daily_pnl_dd <- compute_weights(daily_pos_dd) %>%
  left_join(daily_rets, by = c("date", "ticker")) %>%
  group_by(date) %>%
  summarise(
    portfolio_ret_gross = sum(weight * replace_na(ret_cc, 0), na.rm = TRUE),
    n_long              = first(n_long),
    .groups             = "drop"
  ) %>%
  left_join(rebal_costs_dd, by = "date") %>%
  mutate(
    tc            = replace_na(tc, 0),
    portfolio_ret = portfolio_ret_gross - tc,
    cumeq_dd      = cumprod(1 + portfolio_ret)
  ) %>%
  select(date, cumeq_dd)

# -- Step 5: SPY buy-and-hold benchmark --
spy_bh <- raw_data %>%
  filter(ticker == "SPY", date %in% spy_dates) %>%
  arrange(date) %>%
  mutate(
    ret_cc    = Close / dplyr::lag(Close) - 1,
    cumeq_spy = cumprod(1 + replace_na(ret_cc, 0))
  ) %>%
  select(date, cumeq_spy, ret_cc_spy = ret_cc)

training_perf <- daily_pnl %>%
  left_join(spy_bh,      by = "date") %>%
  left_join(daily_pnl_dd, by = "date")

# Quick summary stats
cat(sprintf(
  "Strategy + DD stop: %.1f%%  |  Strategy (net TC): %.1f%%  |  Strategy (gross): %.1f%%  |  SPY B&H: %.1f%%\nTC drag: %.2f%%\n",
  (last(training_perf$cumeq_dd)    - 1) * 100,
  (last(training_perf$cumeq)       - 1) * 100,
  (last(training_perf$cumeq_gross) - 1) * 100,
  (last(training_perf$cumeq_spy)   - 1) * 100,
  (last(training_perf$cumeq_gross) - last(training_perf$cumeq)) * 100
))
```

## Cumulative Return vs SPY

```{r chart-training-cumeq}
# Rebalancing dates as vertical markers
rebal_markers <- exec_map %>%
  filter(exec_date %in% spy_dates) %>%
  pull(exec_date)

fig_cumeq <- plot_ly(training_perf, x = ~date) %>%
  add_lines(
    y    = ~cumeq_dd,
    name = paste0("Strategy + DD stop (", scales::percent(abs(DD_STOP)), ")"),
    line = list(color = "#59a14f", width = 2)
  ) %>%
  add_lines(
    y    = ~cumeq,
    name = "Strategy (net of TC)",
    line = list(color = "#4e79a7", width = 2)
  ) %>%
  add_lines(
    y    = ~cumeq_gross,
    name = "Strategy (gross)",
    line = list(color = "#4e79a7", width = 1.5, dash = "dot")
  ) %>%
  add_lines(
    y    = ~cumeq_spy,
    name = "SPY Buy & Hold",
    line = list(color = "#e15759", width = 2, dash = "dash")
  ) %>%
  layout(
    title = list(
      text = "Training Period Cumulative Return — Strategy vs SPY Buy & Hold",
      font = list(size = 14)
    ),
    xaxis = list(
      title        = NULL,
      rangeslider  = list(visible = FALSE),
      type         = "date"
    ),
    yaxis = list(
      title      = "Cumulative Return (1.0 = start)",
      tickformat = ".2f"
    ),
    legend     = list(x = 0.01, y = 0.99, bgcolor = "rgba(255,255,255,0.7)"),
    hovermode  = "x unified",
    shapes     = list(
      list(type = "line", x0 = TEST_START, x1 = TEST_START,
           y0 = 0, y1 = 1, yref = "paper",
           line = list(color = "grey50", width = 1, dash = "dot"))
    ),
    annotations = list(
      list(x = TEST_START, y = 1, yref = "paper", xanchor = "left",
           text = " Test start", showarrow = FALSE,
           font = list(size = 10, color = "grey50"))
    )
  )

fig_cumeq
```

## Monthly Return Distribution

```{r chart-monthly-dist}
# Aggregate to monthly returns for both strategy and SPY
monthly_perf <- training_perf %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym) %>%
  summarise(
    strat_monthly = prod(1 + portfolio_ret) - 1,
    spy_monthly   = prod(1 + replace_na(ret_cc_spy, 0)) - 1,
    .groups = "drop"
  ) %>%
  pivot_longer(-ym, names_to = "series", values_to = "monthly_ret") %>%
  mutate(series = recode(series,
    strat_monthly = "Strategy",
    spy_monthly   = "SPY Buy & Hold"
  ))

p_monthly_dist <- ggplot(monthly_perf,
                         aes(x = monthly_ret, fill = series, colour = series)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, alpha = 0.4,
                 position = "identity") +
  geom_density(linewidth = 0.8, alpha = 0) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values   = c("Strategy" = "#4e79a7", "SPY Buy & Hold" = "#e15759")) +
  scale_colour_manual(values = c("Strategy" = "#4e79a7", "SPY Buy & Hold" = "#e15759")) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title    = "Monthly Return Distribution — Training Period",
    subtitle = "Histogram + density overlay  |  Dashed line = 0%",
    x        = "Monthly Return",
    y        = "Density",
    fill     = NULL, colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")

ggplotly(p_monthly_dist, tooltip = c("x", "density"))
```

## Rolling Metrics

```{r chart-drawdown}
dd_data <- training_perf %>%
  arrange(date) %>%
  mutate(
    peak_strat    = cummax(cumeq),
    peak_spy      = cummax(cumeq_spy),
    peak_dd       = cummax(replace_na(cumeq_dd, 1)),
    dd_strat      = (cumeq     - peak_strat) / peak_strat,
    dd_spy        = (cumeq_spy - peak_spy)   / peak_spy,
    dd_strat_stop = (replace_na(cumeq_dd, 1) - peak_dd) / peak_dd
  )

# Max drawdown dates and values for annotations
mdd_strat <- dd_data %>% slice_min(dd_strat, n = 1)
mdd_spy   <- dd_data %>% slice_min(dd_spy,   n = 1)
mdd_dd    <- dd_data %>% slice_min(dd_strat_stop, n = 1)

plot_ly(dd_data, x = ~date) %>%
  add_lines(y = ~dd_strat_stop,
            name = paste0("Strategy + DD stop (", scales::percent(abs(DD_STOP)), ")"),
            line = list(color = "#59a14f", width = 1.5)) %>%
  add_lines(y = ~dd_strat,
            name = "Strategy (net TC)",
            line = list(color = "#4e79a7", width = 2)) %>%
  add_lines(y = ~dd_spy,
            name = "SPY Buy & Hold",
            line = list(color = "#e15759", width = 2, dash = "dash")) %>%
  add_markers(
    data = mdd_strat,
    x = ~date, y = ~dd_strat,
    name = paste0("Strategy MDD: ", scales::percent(mdd_strat$dd_strat, accuracy = 0.1)),
    marker = list(color = "#4e79a7", size = 10, symbol = "circle-open", line = list(width = 2)),
    showlegend = TRUE
  ) %>%
  add_markers(
    data = mdd_spy,
    x = ~date, y = ~dd_spy,
    name = paste0("SPY MDD: ", scales::percent(mdd_spy$dd_spy, accuracy = 0.1)),
    marker = list(color = "#e15759", size = 10, symbol = "circle-open", line = list(width = 2)),
    showlegend = TRUE
  ) %>%
  add_markers(
    data = mdd_dd,
    x = ~date, y = ~dd_strat_stop,
    name = paste0("DD-stop MDD: ", scales::percent(mdd_dd$dd_strat_stop, accuracy = 0.1)),
    marker = list(color = "#59a14f", size = 10, symbol = "circle-open", line = list(width = 2)),
    showlegend = TRUE
  ) %>%
  add_segments(
    x = ~TRAIN_START, xend = ~TRAIN_END,
    y = DD_STOP, yend = DD_STOP,
    line = list(color = "orange", width = 1, dash = "dot"),
    name = paste0("DD stop threshold (", scales::percent(DD_STOP, accuracy = 1), ")"),
    inherit = FALSE
  ) %>%
  layout(
    title     = list(text = "Drawdown from Peak — Training Period", font = list(size = 14)),
    xaxis     = list(title = NULL, type = "date"),
    yaxis     = list(title = "Drawdown from Peak",
                     tickformat = ".0%", zeroline = TRUE, zerolinecolor = "#ccc"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.01, bgcolor = "rgba(255,255,255,0.8)")
  )
```

## Summary Statistics

```{r training-stats}
n_days   <- nrow(training_perf)
n_years  <- n_days / 252

strat_rets       <- training_perf$portfolio_ret
strat_rets_gross <- training_perf$portfolio_ret_gross
spy_rets   <- replace_na(training_perf$ret_cc_spy, 0)

# Join rf_daily for Sharpe
rf_joined <- training_perf %>%
  left_join(rf_daily, by = "date") %>%
  mutate(rf_daily = replace_na(rf_daily, 0))

excess_strat <- strat_rets    - rf_joined$rf_daily
excess_spy   <- spy_rets      - rf_joined$rf_daily

sharpe_strat <- (mean(excess_strat, na.rm = TRUE) /
                 sd(excess_strat,   na.rm = TRUE)) * sqrt(252)
sharpe_spy   <- (mean(excess_spy,   na.rm = TRUE) /
                 sd(excess_spy,     na.rm = TRUE)) * sqrt(252)

total_strat        <- last(training_perf$cumeq)       - 1
total_strat_gross  <- last(training_perf$cumeq_gross) - 1
total_spy          <- last(training_perf$cumeq_spy)   - 1
cagr_strat         <- (1 + total_strat)^(1/n_years)       - 1
cagr_strat_gross   <- (1 + total_strat_gross)^(1/n_years) - 1
cagr_spy           <- (1 + total_spy)^(1/n_years)         - 1
ann_vol_strat      <- sd(strat_rets,       na.rm = TRUE) * sqrt(252)
ann_vol_strat_gross<- sd(strat_rets_gross, na.rm = TRUE) * sqrt(252)
ann_vol_spy        <- sd(spy_rets,         na.rm = TRUE) * sqrt(252)

peak_strat        <- cummax(training_perf$cumeq)
peak_strat_gross  <- cummax(training_perf$cumeq_gross)
peak_spy          <- cummax(training_perf$cumeq_spy)
peak_dd_stop      <- cummax(replace_na(training_perf$cumeq_dd, 1))
max_dd_strat      <- min((training_perf$cumeq       - peak_strat)       / peak_strat,       na.rm = TRUE)
max_dd_strat_gross<- min((training_perf$cumeq_gross - peak_strat_gross) / peak_strat_gross, na.rm = TRUE)
max_dd_spy        <- min((training_perf$cumeq_spy   - peak_spy)         / peak_spy,         na.rm = TRUE)
max_dd_dd_stop    <- min((replace_na(training_perf$cumeq_dd, 1) - peak_dd_stop) / peak_dd_stop, na.rm = TRUE)

# Dates of max drawdown for each series
mdd_date_strat   <- training_perf$date[which.min((training_perf$cumeq - peak_strat) / peak_strat)]
mdd_date_spy     <- training_perf$date[which.min((training_perf$cumeq_spy - peak_spy) / peak_spy)]
mdd_date_dd_stop <- training_perf$date[which.min((replace_na(training_perf$cumeq_dd, 1) - peak_dd_stop) / peak_dd_stop)]

excess_strat_gross <- strat_rets_gross - rf_joined$rf_daily
sharpe_strat_gross <- (mean(excess_strat_gross, na.rm = TRUE) /
                       sd(excess_strat_gross,   na.rm = TRUE)) * sqrt(252)

tibble(
  Metric = c("Total Return", "CAGR", "Ann. Volatility", "Sharpe Ratio",
             "Max Drawdown", "MDD Date", "TC Drag", "Period"),
  `Strategy + DD Stop` = c(
    scales::percent(last(training_perf$cumeq_dd, na.rm = TRUE) - 1, accuracy = 0.1),
    scales::percent((last(training_perf$cumeq_dd, na.rm = TRUE))^(1/n_years) - 1, accuracy = 0.1),
    "—",
    "—",
    scales::percent(max_dd_dd_stop, accuracy = 0.1),
    format(mdd_date_dd_stop, "%Y-%m-%d"),
    "—",
    paste(format(TRAIN_START, "%Y-%m-%d"), "–", format(TRAIN_END, "%Y-%m-%d"))
  ),
  `Strategy (net TC)` = c(
    scales::percent(total_strat,       accuracy = 0.1),
    scales::percent(cagr_strat,        accuracy = 0.1),
    scales::percent(ann_vol_strat,     accuracy = 0.1),
    round(sharpe_strat, 3),
    scales::percent(max_dd_strat,      accuracy = 0.1),
    format(mdd_date_strat, "%Y-%m-%d"),
    scales::percent(total_strat_gross - total_strat, accuracy = 0.1),
    paste(format(TRAIN_START, "%Y-%m-%d"), "–", format(TRAIN_END, "%Y-%m-%d"))
  ),
  `Strategy (gross)` = c(
    scales::percent(total_strat_gross,      accuracy = 0.1),
    scales::percent(cagr_strat_gross,       accuracy = 0.1),
    scales::percent(ann_vol_strat_gross,    accuracy = 0.1),
    round(sharpe_strat_gross, 3),
    scales::percent(max_dd_strat_gross,     accuracy = 0.1),
    "—",
    "—",
    paste(format(TRAIN_START, "%Y-%m-%d"), "–", format(TRAIN_END, "%Y-%m-%d"))
  ),
  `SPY Buy & Hold` = c(
    scales::percent(total_spy,    accuracy = 0.1),
    scales::percent(cagr_spy,     accuracy = 0.1),
    scales::percent(ann_vol_spy,  accuracy = 0.1),
    round(sharpe_spy, 3),
    scales::percent(max_dd_spy,   accuracy = 0.1),
    format(mdd_date_spy, "%Y-%m-%d"),
    "—",
    paste(format(TRAIN_START, "%Y-%m-%d"), "–", format(TRAIN_END, "%Y-%m-%d"))
  )
) %>%
  knitr::kable(caption = "Training Period Performance Summary (5 bps/side TC)", align = "lrrrr")
```
