---
title: "FIN 452 — Quantitative Trading Strategy"
subtitle: "Regime-Based Sector Rotation via Clustering & Kalman Filter"
author: "Bao Nguyen"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-width: 12
    fig-height: 8
execute:
  warning: false
  message: false
  echo: true
  cache: false
---

```{r setup}
library(tidyverse)
library(tidyquant)
library(tidymodels)
library(tidyclust)
library(timetk)
library(TTR)
library(PerformanceAnalytics)
library(xts)
library(zoo)
library(quantmod)
library(RTL)
library(lattice)
library(plotly)
library(foreach)
library(doParallel)
library(dlm)
library(cluster)
library(factoextra)

# Helper: rolling function with partial window support
# Computes as soon as min_obs are available; doesn't wait for full window.
# Allows first-month ETF data to produce values without the full lookback.
roll_partial <- function(x, width, FUN, min_obs = 5, ...) {
  rollapply(x, width = width, FUN = function(z) {
    z_clean <- z[!is.na(z)]
    if (length(z_clean) < min_obs) return(NA_real_)
    FUN(z_clean, ...)
  }, fill = NA, align = "right", partial = TRUE)
}

# Helper: rolling beta (two-series covariance / variance)
roll_beta <- function(sector_ret, spy_ret, width = 252, min_obs = 20) {
  mat <- cbind(sector_ret, spy_ret)
  rollapply(mat, width = width, FUN = function(x) {
    cc <- complete.cases(x)
    if (sum(cc) < min_obs) return(NA_real_)
    cov(x[cc, 1], x[cc, 2]) / var(x[cc, 2])
  }, by.column = FALSE, fill = NA, align = "right", partial = TRUE)
}
```

## Universe & Periods

```{r params}
tickers <- c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
             "XLK", "XLP", "XLRE", "XLU", "XLV", "XLY")

# Availability constraints
etf_inception <- tribble(
  ~ticker, ~inception,
  "SPY",   "1993-01-29",
  "XLB",   "1998-12-22",
  "XLC",   "2018-06-18",
  "XLE",   "1998-12-22",
  "XLF",   "1998-12-22",
  "XLI",   "1998-12-22",
  "XLK",   "1998-12-22",
  "XLP",   "1998-12-22",
  "XLRE",  "2015-10-07",
  "XLU",   "1998-12-22",
  "XLV",   "1998-12-22",
  "XLY",   "1998-12-22"
) %>% mutate(inception = as.Date(inception))

TRAIN_START <- as.Date("2000-06-01")
TRAIN_END   <- as.Date("2024-12-31")
TEST_START  <- as.Date("2025-01-01")
TEST_END    <- as.Date("2026-03-31")

PULL_FROM   <- as.Date("1998-01-01")   # pull full history; filter later
```

## Data Loading

```{r load-data}
#| cache: true

# ⚠️ COURSE PATTERN: tq_get → toTitleCase → xts → adjustOHLC → tibble
load_etf <- function(ticker) {
  tidyquant::tq_get(ticker, from = format(PULL_FROM)) %>%
    dplyr::rename_all(tools::toTitleCase) %>%
    timetk::tk_xts(date_var = Date) %>%
    quantmod::adjustOHLC(use.Adjusted = TRUE) %>%
    timetk::tk_tbl(rename_index = "Date") %>%
    dplyr::select(-Adjusted) %>%
    dplyr::mutate(across(where(is.numeric), round, 2)) %>%
    dplyr::rename(date = Date) %>%
    dplyr::mutate(ticker = ticker)
}

raw_data <- tickers %>%
  purrr::map_dfr(load_etf) %>%
  left_join(etf_inception, by = "ticker") %>%
  filter(date >= inception) %>%        # drop pre-inception rows
  select(-inception) %>%
  arrange(ticker, date)

# Training and test subsets
train_data <- raw_data %>% filter(date >= TRAIN_START, date <= TRAIN_END)
test_data  <- raw_data %>% filter(date >= TEST_START,  date <= TEST_END)

glimpse(raw_data)
```

## Interest Rates

10Y Treasury yield (`^TNX`) from Yahoo Finance, 2Y Treasury yield (`DGS2`) from FRED. The yield curve slope uses 10Y − 2Y — more responsive to rate cycle turning points than 10Y − 3M because the 2Y prices in Fed expectations before they act, which is what moves rate-sensitive sectors (XLF, XLRE, XLU).

```{r load-rates}
#| cache: true

rates_raw <- bind_rows(
  tq_get("^TNX", from = format(PULL_FROM)) %>%
    transmute(date, series = "y10", value = close),
  fredr::fredr("DGS2", observation_start = PULL_FROM) %>%
    transmute(date, series = "y2", value)
) %>%
  pivot_wider(names_from = series, values_from = value)

# Yield curve slope: 10Y minus 2Y (percentage points)
# Positive = normal curve; negative = inverted (recession signal)
rates <- rates_raw %>%
  mutate(yield_curve = y10 - y2) %>%
  filter(date >= TRAIN_START)
```

```{r chart-rates}
rates_long <- rates %>%
  pivot_longer(-date, names_to = "series", values_to = "value") %>%
  mutate(series = recode(series,
    y10 = "10Y Treasury", y2 = "2Y Treasury", yield_curve = "Yield Curve (10Y − 2Y)"))

p_rates <- ggplot(rates_long, aes(x = date, y = value, colour = series)) +

  annotate("rect", xmin = TEST_START,
           xmax = max(rates_long$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +

  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.4) +

  geom_hline(yintercept = 0, linetype = "dotted",
             colour = "grey50", linewidth = 0.3) +

  geom_line(linewidth = 0.5, alpha = 0.85) +

  scale_colour_manual(values = c(
    "10Y Treasury"            = "#4e79a7",
    "3M T-Bill"               = "#76b7b2",
    "Yield Curve (10Y − 3M)"  = "#e15759"
  )) +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +

  labs(
    title    = "Treasury Yields & Yield Curve Slope",
    subtitle = "Negative yield curve (red below zero) = inverted = recession signal",
    x = NULL, y = "Yield (%)", colour = NULL
  ) +

  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))

ggplotly(p_rates, tooltip = c("x", "y", "colour")) %>%
  layout(legend = list(orientation = "h", y = 1.08))
```

## Historical Price Levels

Each ETF indexed to 100 at its first available date within the chart window. Vertical line marks the train/test split. Greyed region = out-of-sample test period.

```{r chart-levels}
# Index each ETF to 100 at its first observation, then pivot wide (one column per ticker)
display_data <- raw_data %>%
  filter(date >= TRAIN_START) %>%
  group_by(ticker) %>%
  arrange(date) %>%
  mutate(level = 100 * Close / first(Close)) %>%
  ungroup() %>%
  select(date, ticker, level) %>%
  pivot_wider(names_from = ticker, values_from = level)

# Pivot back to long for plotting
p <- display_data %>%
  pivot_longer(-date, names_to = "ticker", values_to = "level") %>%
  mutate(ticker = factor(ticker, levels = tickers)) %>%
  ggplot(aes(x = date, y = level)) +

  # Test period shading
  annotate("rect",
           xmin = TEST_START, xmax = TEST_END,
           ymin = -Inf, ymax = Inf,
           fill = "#f0f0f0", alpha = 0.8) +

  # Train/test split line
  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.4) +

  geom_line(colour = "steelblue", linewidth = 0.4) +

  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +

  annotate("text",
           x = TEST_START + 30, y = -Inf,
           label = "Test", hjust = 0, vjust = -0.5,
           size = 2.5, colour = "grey40") +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(suffix = "")) +

  labs(
    title    = "SPDR Sector ETF Historical Levels (Indexed to 100 at Training Start)",
    subtitle = "Training: Jun 2000 – Dec 2024  |  Test: Jan 2025 – Mar 2026  |  * = later inception date",
    x        = NULL,
    y        = "Indexed Level (100 = first observation)"
  ) +

  theme_minimal(base_size = 10) +
  theme(
    strip.text       = element_text(size = 7.5, face = "bold"),
    axis.text.x      = element_text(size = 7, angle = 30, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )

ggplotly(p, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

## Rolling Volatility

### Compute

```{r compute-features}
#| cache: true

# Pull VIX — already annualised % (e.g. 20 = 20% annual vol), no transformation needed
vix_raw <- tidyquant::tq_get("^VIX", from = format(PULL_FROM)) %>%
  select(date, vix = close)

# Get SPY returns separately for beta computation
spy_returns <- raw_data %>%
  filter(ticker == "SPY", date >= TRAIN_START) %>%
  arrange(date) %>%
  mutate(spy_retClCl = Close / lag(Close) - 1,
         spy_log_ret = log(Close / lag(Close))) %>%
  select(date, spy_retClCl, spy_log_ret)

# 21 trading days = 30 calendar days * 252/365 ≈ 21
# This matches VIX's 30-calendar-day horizon for apples-to-apples VRP
RVOL_WINDOW  <- 21
BETA_WINDOW  <- 252
ZSCORE_WINDOW <- 252

# Compute all rolling features per ETF
# partial = TRUE via roll_partial() allows first-month data to produce values
features_long <- raw_data %>%
  filter(date >= TRAIN_START) %>%
  group_by(ticker) %>%
  arrange(date) %>%
  mutate(
    retClCl     = Close / lag(Close) - 1,
    log_ret     = log(Close / lag(Close)),

    # 21d rolling vol (annualised %)
    rvol_21     = roll_partial(retClCl, width = RVOL_WINDOW, FUN = sd,
                               min_obs = 5) * sqrt(252) * 100,

    # 21d rolling log return (sum of daily log returns)
    # Partial window: first-month ETFs get whatever days are available
    log_ret_21  = roll_partial(log_ret, width = RVOL_WINDOW, FUN = sum,
                               min_obs = 1)
  ) %>%
  ungroup() %>%
  # Join SPY returns for beta computation
  left_join(spy_returns, by = "date") %>%
  # Join VIX for spread and VRP
  left_join(vix_raw, by = "date") %>%
  group_by(ticker) %>%
  arrange(date) %>%
  mutate(
    # 252d rolling beta to SPY (partial: starts computing at 20 obs)
    beta_252 = roll_beta(retClCl, spy_retClCl, width = BETA_WINDOW, min_obs = 20),

    # Vol spread: sector realised vol minus VIX
    # Levels are biased by sector character — z-score removes this
    spread   = rvol_21 - vix,

    # 252d rolling z-score of the spread
    # Removes sector-specific mean AND variance → directly comparable across sectors
    spread_z = {
      sp_mean <- roll_partial(spread, width = ZSCORE_WINDOW, FUN = mean, min_obs = 30)
      sp_sd   <- roll_partial(spread, width = ZSCORE_WINDOW, FUN = sd,   min_obs = 30)
      (spread - sp_mean) / sp_sd
    }
  ) %>%
  ungroup() %>%
  select(date, ticker, retClCl, log_ret, rvol_21, log_ret_21,
         beta_252, vix, spread, spread_z)

# Wide format: one row per date, metrics as columns per ticker
features_wide <- features_long %>%
  select(date, ticker, rvol_21, spread, spread_z, log_ret_21, beta_252) %>%
  pivot_wider(names_from = ticker,
              values_from = c(rvol_21, spread, spread_z, log_ret_21, beta_252),
              names_glue = "{ticker}_{.value}") %>%
  left_join(vix_raw, by = "date") %>%
  left_join(rates %>% select(date, yield_curve, y10, y2), by = "date") %>%
  mutate(vrp = vix - SPY_rvol_21)   # VRP: apples-to-apples, both 21 trading days
```

### SPY Rolling Vol vs VIX

21-trading-day realised vol for SPY plotted against VIX. Both cover ~30 calendar days (21 trading days = 30 * 252/365) — directly comparable. The gap is the Volatility Risk Premium (VRP): positive = fear premium; negative = crash worse than anticipated.

```{r chart-spy-vol}
spy_vol <- features_wide %>%
  select(date, `21d Realised` = SPY_rvol_21, VIX = vix, VRP = vrp) %>%
  pivot_longer(-date, names_to = "series", values_to = "vol") %>%
  mutate(series = factor(series, levels = c("VIX", "21d Realised", "VRP")))

p_spy <- ggplot(spy_vol, aes(x = date, y = vol, colour = series)) +

  annotate("rect",
           xmin = TEST_START, xmax = max(spy_vol$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +

  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.4) +

  geom_hline(yintercept = 0, linetype = "dotted",
             colour = "grey60", linewidth = 0.3) +

  geom_line(linewidth = 0.5, alpha = 0.85) +

  scale_colour_manual(values = c(
    "VIX"          = "#e15759",
    "21d Realised" = "#4e79a7",
    "VRP"          = "#f28e2b"
  )) +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +

  labs(
    title    = "SPY: VIX vs 21d Realised Volatility & VRP",
    subtitle = "VRP = VIX − 21d Realised Vol (21 trading days ≈ 30 calendar days)  |  Positive = fear premium",
    x        = NULL, y = "Annualised Volatility (%)", colour = NULL
  ) +

  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "top",
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )

ggplotly(p_spy, tooltip = c("x", "y", "colour")) %>%
  layout(legend = list(orientation = "h", y = 1.08))
```

### Sector ETF Vol Spread Z-Score

Rolling z-score of each sector's vol spread (sector rvol − VIX), normalised by that sector's own 252-day rolling mean and standard deviation. Removes structural level differences — a z-score of +2 means the same anomalous stress regardless of whether the sector is normally above or below VIX. Partial window support: values appear earlier than 252 days but require at least 30 observations for the z-score.

```{r chart-sector-vol}
sector_tickers <- setdiff(tickers, "SPY")

sector_vol <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  mutate(ticker = factor(ticker, levels = sector_tickers))

p_sec <- ggplot(sector_vol, aes(x = date, y = spread_z)) +

  annotate("rect",
           xmin = TEST_START, xmax = max(sector_vol$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +

  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.3) +

  # Zero line = sector's historical norm
  geom_hline(yintercept = 0, linetype = "dotted",
             colour = "grey50", linewidth = 0.3) +

  geom_line(colour = "#4e79a7", linewidth = 0.35, alpha = 0.85) +

  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +

  labs(
    title    = "Sector Vol Spread Z-Score  (sector rvol − VIX, normalised)",
    subtitle = "Zero = historical norm  |  +2 = 2sd more stressed vs VIX than usual  |  Comparable across sectors",
    x        = NULL, y = "Z-Score (standard deviations)"
  ) +

  theme_minimal(base_size = 10) +
  theme(
    strip.text       = element_text(size = 8, face = "bold"),
    axis.text.x      = element_text(size = 7, angle = 30, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )

ggplotly(p_sec, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

## Rolling Beta to SPY

252-day rolling beta per sector vs SPY. Measures each sector's market sensitivity over time — a regime signal. When beta changes, the sector is coupling to or decoupling from the broad market. Partial window: starts producing values at 20 observations.

```{r chart-beta}
beta_data <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  mutate(ticker = factor(ticker, levels = sector_tickers))

p_beta <- ggplot(beta_data, aes(x = date, y = beta_252)) +

  annotate("rect",
           xmin = TEST_START, xmax = max(beta_data$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +

  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.3) +

  geom_hline(yintercept = 1, linetype = "dotted",
             colour = "grey50", linewidth = 0.3) +

  geom_line(colour = "#59a14f", linewidth = 0.35, alpha = 0.85) +

  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +

  labs(
    title    = "252d Rolling Beta to SPY",
    subtitle = "Beta = 1 line shown  |  Above 1 = amplifies SPY moves  |  Below 1 = dampens / decouples",
    x        = NULL, y = "Beta"
  ) +

  theme_minimal(base_size = 10) +
  theme(
    strip.text       = element_text(size = 8, face = "bold"),
    axis.text.x      = element_text(size = 7, angle = 30, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )

ggplotly(p_beta, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```

## 21d Rolling Log Return

Cumulative log return over 21 trading days (~1 month). Captures short-term sector momentum. Partial window: first-month ETFs produce values from whatever days are available.

```{r chart-logret}
logret_data <- features_long %>%
  filter(ticker %in% sector_tickers) %>%
  mutate(ticker = factor(ticker, levels = sector_tickers))

p_logret <- ggplot(logret_data, aes(x = date, y = log_ret_21)) +

  annotate("rect",
           xmin = TEST_START, xmax = max(logret_data$date, na.rm = TRUE),
           ymin = -Inf, ymax = Inf, fill = "#f0f0f0", alpha = 0.8) +

  geom_vline(xintercept = TEST_START, linetype = "dashed",
             colour = "grey40", linewidth = 0.3) +

  geom_hline(yintercept = 0, linetype = "dotted",
             colour = "grey50", linewidth = 0.3) +

  geom_line(colour = "#b07aa1", linewidth = 0.35, alpha = 0.85) +

  facet_wrap(~ ticker, scales = "free_y", ncol = 3) +

  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_percent()) +

  labs(
    title    = "21d Rolling Log Return (Sector Momentum)",
    subtitle = "Cumulative log return over 21 trading days  |  Positive = upward momentum  |  Negative = downward",
    x        = NULL, y = "Log Return"
  ) +

  theme_minimal(base_size = 10) +
  theme(
    strip.text       = element_text(size = 8, face = "bold"),
    axis.text.x      = element_text(size = 7, angle = 30, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )

ggplotly(p_logret, tooltip = c("x", "y")) %>%
  layout(showlegend = FALSE)
```
